### 2025-06-18 paper.tex
check

%%
%% This is file `sample-sigplan.tex',
%% generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `all,proceedings,bibtex,sigplan')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed
%% with new filenames distinct from sample-sigplan.tex.
%% 
%% For distribution of the original source see the terms
%% for copying and modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the
%% original source files, as listed above, are part of the
%% same distribution. (The sources need not necessarily be
%% in the same archive or directory.)
%%
%%
%% Commands for TeXCount
%TC:macro \cite [option:text,text]
%TC:macro \citep [option:text,text]
%TC:macro \citet [option:text,text]
%TC:envir table 0 1
%TC:envir table* 0 1
%TC:envir tabular [ignore] word
%TC:envir displaymath 0 word
%TC:envir math 0 word
%TC:envir comment 0 0
%%
%% The first command in your LaTeX source must be the \documentclass
%% command.
%%
%% For submission and review of your manuscript please change the
%% command to \documentclass[manuscript, screen, review]{acmart}.
%%
%% When submitting camera ready or to TAPS, please change the command
%% to \documentclass[sigconf]{acmart} or whichever template is required
%% for your publication.
%%
%%
\documentclass[sigplan,review,dvipsnames,screen,10pt]{acmart}
%\documentclass[sigplan,screen]{acmart}
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    Bib\TeX}}}

%% Rights management information.  This information is sent to you
%% when you complete the rights form.  These commands have SAMPLE
%% values in them; it is your responsibility as an author to replace
%% the commands and values with those provided to you when you
%% complete the rights form.
\setcopyright{acmlicensed}
\copyrightyear{2018}
\acmYear{2018}
\acmDOI{XXXXXXX.XXXXXXX}
%% These commands are for a PROCEEDINGS abstract or paper.
\acmConference[OlivierFest '25]{Workshop dedicated to Olivier Danvy's 65th birthday}{October,
  2025}{Singapore}
%%
%%  Uncomment \acmBooktitle if the title of the proceedings is different
%%  from ``Proceedings of ...''!
%%
%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
%%  June 03--05, 2018, Woodstock, NY}
\acmISBN{978-1-4503-XXXX-X/2018/06}

%%
%% Submission ID.
%% Use this when submitting an article to a sponsored event. You'll
%% receive a unique submission ID from the organizers
%% of the event, and this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% For managing citations, it is recommended to use bibliography
%% files in BibTeX format.
%%
%% You can then either use BibTeX with the ACM-Reference-Format style,
%% or BibLaTeX with the acmnumeric or acmauthoryear sytles, that include
%% support for advanced citation of software artefact from the
%% biblatex-software package, also separately available on CTAN.
%%
%% Look at the sample-*-biblatex.tex files for templates showcasing
%% the biblatex styles.
%%

%%
%% The majority of ACM publications use numbered citations and
%% references.  The command \citestyle{authoryear} switches to the
%% "author year" style.
%%
%% If you are preparing content for an event
%% sponsored by ACM SIGGRAPH, you must use the "author year" style of
%% citations and references.
%% Uncommenting
%% the next command will enable that style.
%%\citestyle{acmauthoryear}

\newcommand{\TRUE}[0]{\mathsf{true}}
\newcommand{\FALSE}[0]{\mathsf{false}}
\newcommand{\IF}[3]{\mathsf{if}\,#1\,\mathsf{then}\,#2\,\mathsf{else}\,#3}

\newcommand{\CE}[1]{\mathcal{E} \llbracket #1 \rrbracket}
\newcommand{\CB}[3]{\mathcal{B} \llbracket #1 \rrbracket ( #2, #3 )}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title{Verified Nanopasses for Compiling Conditionals}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
%% Of note is the shared affiliation of the first two authors, and the
%% "authornote" and "authornotemark" commands
%% used to denote shared contribution to the research.
\author{Jeremy G. Siek}
%\authornote{Both authors contributed equally to this research.}
\email{jsiek@iu.edu}
%\orcid{1234-5678-9012}
%% \author{G.K.M. Tobin}
%% \authornotemark[1]
%% \email{webmaster@marysville-ohio.com}
\affiliation{%
  \institution{Indiana University}
  \city{Bloomington}
  \state{IN}
  \country{USA}
}

%% \author{Lars Th{\o}rv{\"a}ld}
%% \affiliation{%
%%   \institution{The Th{\o}rv{\"a}ld Group}
%%   \city{Hekla}
%%   \country{Iceland}}
%% \email{larst@affiliation.org}

%% \author{Valerie B\'eranger}
%% \affiliation{%
%%   \institution{Inria Paris-Rocquencourt}
%%   \city{Rocquencourt}
%%   \country{France}
%% }

%% \author{Aparna Patel}
%% \affiliation{%
%%  \institution{Rajiv Gandhi University}
%%  \city{Doimukh}
%%  \state{Arunachal Pradesh}
%%  \country{India}}

%% \author{Huifen Chan}
%% \affiliation{%
%%   \institution{Tsinghua University}
%%   \city{Haidian Qu}
%%   \state{Beijing Shi}
%%   \country{China}}

%% \author{Charles Palmer}
%% \affiliation{%
%%   \institution{Palmer Research Laboratories}
%%   \city{San Antonio}
%%   \state{Texas}
%%   \country{USA}}
%% \email{cpalmer@prl.com}

%% \author{John Smith}
%% \affiliation{%
%%   \institution{The Th{\o}rv{\"a}ld Group}
%%   \city{Hekla}
%%   \country{Iceland}}
%% \email{jsmith@affiliation.org}

%% \author{Julius P. Kumquat}
%% \affiliation{%
%%   \institution{The Kumquat Consortium}
%%   \city{New York}
%%   \country{USA}}
%% \email{jpkumquat@consortium.net}

%%
%% By default, the full list of authors will be used in the page
%% headers. Often, this list is too long, and will overlap
%% other information printed in the page headers. This command allows
%% the author to define a more concise list
%% of authors' names for this purpose.
%\renewcommand{\shortauthors}{Trovato et al.}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
We present a proof of correctness in Agda for four nanopasses that
translate a source language with let binding, integer arithmetic,
conditional expressions and Booleans into an intermediate language
that is roughly x86 assembly with variables (prior to register
allocation).  The most interesting of these four nanopasses is a
translation of conditional expressions into goto-style control flow
that uses the continuation-oriented approach of Olivier Danvy's
one-pass transformation into monadic normal form (2003).
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm.
%% Please copy and paste the code instead of the example below.
%%
%% \begin{CCSXML}
%% <ccs2012>
%%  <concept>
%%   <concept_id>00000000.0000000.0000000</concept_id>
%%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
%%   <concept_significance>500</concept_significance>
%%  </concept>
%%  <concept>
%%   <concept_id>00000000.00000000.00000000</concept_id>
%%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
%%   <concept_significance>300</concept_significance>
%%  </concept>
%%  <concept>
%%   <concept_id>00000000.00000000.00000000</concept_id>
%%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
%%   <concept_significance>100</concept_significance>
%%  </concept>
%%  <concept>
%%   <concept_id>00000000.00000000.00000000</concept_id>
%%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
%%   <concept_significance>100</concept_significance>
%%  </concept>
%% </ccs2012>
%% \end{CCSXML}

%% \ccsdesc[500]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
%% \ccsdesc[300]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
%% \ccsdesc{Do Not Use This Code~Generate the Correct Terms for Your Paper}
%% \ccsdesc[100]{Do Not Use This Code~Generate the Correct Terms for Your Paper}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{short-cut boolean evaluation, nanopass, verified compilation, mechanized proof, Agda}
%% A "teaser" image appears between the author and affiliation
%% information and the body of the document, and typically spans the
%% page.
%% \begin{teaserfigure}
%%   \includegraphics[width=\textwidth]{sampleteaser}
%%   \caption{Seattle Mariners at Spring Training, 2010.}
%%   \Description{Enjoying the baseball game from the third-base
%%   seats. Ichiro Suzuki preparing to bat.}
%%   \label{fig:teaser}
%% \end{teaserfigure}

%% \received{20 February 2007}
%% \received[revised]{12 March 2009}
%% \received[accepted]{5 June 2009}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\section{Introduction}

The compilation of Boolean expressions and conditionals is a classic
topic in the compiler literature, going back to a sophisticated
algorithm of \citet{Aho:1986qf} (presented in the form of an attribute
grammar) that translates Boolean expressions into gotos. The
compilation of Boolean expressions and conditionals often occurs at an
interesting stage in the compilation pipeline, where the intermediate
language changes from an abstract syntax \emph{tree} into a control
flow \emph{graph}, or equivalently, an intermediate language with
\texttt{goto}. Further, when targeting an assembly language such as
Intel's x86, the naive approach to compiling boolean expressions and
conditionals produces embarassingly inefficient code. Similar issues
arise in the compilation of languages such as Haskell, where the
case-of-case transformation is employed to optimize dispatching on
algebraic datatypes \citep{PEYTONJONES19983}.

Olivier Danvy, in his paper \emph{A New One-Pass Transformation into
Monadic Normal Form}~\citep{Danvy:2003fk}, presents a beautiful family
of structurally recursive functions that translate Boolean expressions
and conditionals into a monadic normal form that no longer contains
Boolean expressions or nested conditionals, and where jumps are
expressed using zero-arity function calls. One of the key parts of
that translation is in the compilation of a conditional expression
$(\IF{b}{e_1}{e_2})$, where the expression $b$ is compiled using a
special function $\mathcal{B}$ that takes two more arguments, the
generated code for the two branches. Thus, $\mathcal{B}$ can analyze
the expression $b$ to decide how to generate code for the
$\mathsf{if}$, or whether it is even necessary.
\[
\CE{\IF{b}{e_1}{e_2}} = \CB{b}{\CE{e_1}}{\CE{e_2}}
\]
For example, in the simple cases where $b$ is $\TRUE$ or $\FALSE$,
no branching is needed.
\begin{align*}
\CB{\TRUE}{c_1}{c_2} &= c_1 \\
\CB{\FALSE}{c_1}{c_2} &= c_2
\end{align*}

My first exposure to these ideas were in Kent Dybvig's compiler
course~\citep{Dybvig:2010aa}, in which he taught the techniques that
he employed in building Chez Scheme~\citep{Dybvig:2006aa}, including
destintation-driven code generation~\cite{Dybvig:1990aa}. The idea
there was to use an extra parameter in the compilation function to
indicate how the context would use the result value of an
expression. The context might (1) ignore the result (and only care
about the expression's side effects), (2) branch based on interpreting
its result as a boolean, or (3) actually use the result
value. Relating this to Olivier's transformation functions, the
$\mathcal{E}$ function is for contexts that need the actual value and
the the $\mathcal{B}$ function is for contexts that branch on the
expression's result. Of course, a pair of function is isomorphic to a
function on a disjoint union.
\[
(A \to B) \times (C \to B) \cong (A + C) \to B
\]

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{all}

%%
%% If your work has an appendix, this is the place to put it.
%\appendix

\end{document}
\endinput
%%
%% End of file `sample-sigplan.tex'.

### 2025-06-18 paper.tex
tweak

@@
 translate a source language with let binding, integer arithmetic,
-conditional expressions and Booleans into an intermediate language
-that is roughly x86 assembly with variables (prior to register
-allocation).  The most interesting of these four nanopasses is a
+conditional expressions and Booleans into an x86-flavored register
+transfer language. The most interesting of these four nanopasses is a
 translation of conditional expressions into goto-style control flow
@@
 The compilation of Boolean expressions and conditionals is a classic
-topic in the compiler literature, going back to a sophisticated
-algorithm of \citet{Aho:1986qf} (presented in the form of an attribute
-grammar) that translates Boolean expressions into gotos. The
+topic in the compiler literature, going back to an algorithm of
+\citet{Aho:1986qf} (presented in the form of an attribute grammar)
+that translates Boolean expressions and conditionals into gotos. The
 compilation of Boolean expressions and conditionals often occurs at an
@@
 that translation is in the compilation of a conditional expression
-$(\IF{b}{e_1}{e_2})$, where the expression $b$ is compiled using a
-special function $\mathcal{B}$ that takes two more arguments, the
-generated code for the two branches. Thus, $\mathcal{B}$ can analyze
-the expression $b$ to decide how to generate code for the
-$\mathsf{if}$, or whether it is even necessary.
+such as $(\IF{b}{e_1}{e_2})$, where the expression $b$ is compiled
+using a special function $\mathcal{B}$ that takes two more arguments,
+the generated code for the two branches. Thus, $\mathcal{B}$ can
+analyze the expression $b$ to decide how to generate code for the
+$\mathsf{if}$, or whether an $\mathsf{if}$ is even necessary.
 \[
@@
 My first exposure to these ideas were in Kent Dybvig's compiler
-course~\citep{Dybvig:2010aa}, in which he taught the techniques that
-he employed in building Chez Scheme~\citep{Dybvig:2006aa}, including
-destintation-driven code generation~\cite{Dybvig:1990aa}. The idea
-there was to use an extra parameter in the compilation function to
-indicate how the context would use the result value of an
-expression. The context might (1) ignore the result (and only care
-about the expression's side effects), (2) branch based on interpreting
-its result as a boolean, or (3) actually use the result
-value. Relating this to Olivier's transformation functions, the
-$\mathcal{E}$ function is for contexts that need the actual value and
-the the $\mathcal{B}$ function is for contexts that branch on the
-expression's result. Of course, a pair of function is isomorphic to a
-function on a disjoint union.
+course~\citep{Dybvig:2010aa} where he taught the techniques used in
+the Chez Scheme compiler~\citep{Dybvig:2006aa}.  One of those
+techniques is called destintation-driven code
+generation~\cite{Dybvig:1990aa}, which employs an extra parameter on
+the compilation function that indicates how the context will use the
+result value of an expression. The context might (1) ignore the result
+(and only care about the expression's side effects), (2) branch based
+on interpreting its result as a boolean, or (3) actually use the
+result value. These options are encoded as a disjoint union.  Relating
+this to Olivier's transformation functions, the $\mathcal{E}$ function
+is for contexts that need the actual value and the $\mathcal{B}$
+function is for contexts that branch on the expression's result. Of
+course, a pair of functions is isomorphic to a function on a disjoint
+union.
 \[
-(A \to B) \times (C \to B) \cong (A + C) \to B
+(A \to C) \times (B \to C) \cong (A + B) \to C
 \]
+
+The CompCert compiler \citep{Leroy:2006fe}, a compiler for C verified
+in Coq, uses the one-function per kind of context approach, in the
+RTLgen pass that translates from the CminorSel intermediate language
+to its register transfer language.
 

### 2025-06-18 paper.tex
progress

@@
 of structurally recursive functions that translate Boolean expressions
-and conditionals into a monadic normal form that no longer contains
-Boolean expressions or nested conditionals, and where jumps are
-expressed using zero-arity function calls. One of the key parts of
-that translation is in the compilation of a conditional expression
-such as $(\IF{b}{e_1}{e_2})$, where the expression $b$ is compiled
-using a special function $\mathcal{B}$ that takes two more arguments,
-the generated code for the two branches. Thus, $\mathcal{B}$ can
-analyze the expression $b$ to decide how to generate code for the
-$\mathsf{if}$, or whether an $\mathsf{if}$ is even necessary.
+and conditionals into a monadic normal form that (1) no longer
+contains Boolean expressions, (2) the condition expression of a
+condition is a value, and (3) jumps are expressed using zero-arity
+function calls. One of the key parts of that translation is in the
+compilation of a conditional expression such as $(\IF{b}{e_1}{e_2})$,
+where the expression $b$ is compiled using a special function
+$\mathcal{B}$ that takes two more arguments, the generated code for
+the two branches. Thus, $\mathcal{B}$ can analyze the expression $b$
+to decide how to generate code for the $\mathsf{if}$, or whether an
+$\mathsf{if}$ is even necessary.
 \[
@@
 techniques is called destintation-driven code
-generation~\cite{Dybvig:1990aa}, which employs an extra parameter on
-the compilation function that indicates how the context will use the
-result value of an expression. The context might (1) ignore the result
-(and only care about the expression's side effects), (2) branch based
-on interpreting its result as a boolean, or (3) actually use the
-result value. These options are encoded as a disjoint union.  Relating
-this to Olivier's transformation functions, the $\mathcal{E}$ function
-is for contexts that need the actual value and the $\mathcal{B}$
-function is for contexts that branch on the expression's result. Of
-course, a pair of functions is isomorphic to a function on a disjoint
-union.
+generation~\cite{Dybvig:1990aa}, which adds an extra parameter to the
+compilation function to indicate how the context will use the result
+value of an expression. The context may (1) ignore the result (and
+only care about the expression's side effects), (2) branch based on
+interpreting its result as a boolean, (3) assign the result to a
+variable, or (4) output the result. These options are encoded as a
+disjoint union.  Relating this to Olivier's transformation functions,
+the $\mathcal{E}$ function is for contexts that need the actual value
+and the $\mathcal{B}$ function is for contexts that branch on the
+expression's result. Of course, a pair of functions is isomorphic to a
+function on a disjoint union.
 \[
@@
 
-The CompCert compiler \citep{Leroy:2006fe}, a compiler for C verified
-in Coq, uses the one-function per kind of context approach, in the
+The CompCert compiler \citep{Leroy:2006fe} (a compiler for C verified
+in Coq) uses the one-function per context approach of Danvy in the
 RTLgen pass that translates from the CminorSel intermediate language
-to its register transfer language.
+to CompCert's register transfer language. The verified CakeML
+compiler, on the other hand, employs a naive translation of Boolean
+expressions and conditionals for the compilation of stackLang the
+intermediate language into labLang~\citep{Kumar:2014aa}.
+
+Towards developing a course on verified compilation, I am interested
+in adapting verification techniques from CompCert to the compiler
+presented in the \emph{Essentials of Compilation}
+textbook~\citep{Siek:2023tr,Siek:2023ue}. That book takes an
+incremental approach to teaching compilers, that is, it guides the
+students through the creation of multiple compilers for a sequence of
+languages, each of which adds a language feature to the previous
+langauge. For the language with Boolean expressions and conditionals,
+the Explicate Control pass of the compiler uses the one-function per
+context approach of Danvy to translate into an x86-flavored register
+transfer language.  \emph{Essentials of Compilation} also employs the
+nanopass approach, breaking up the compiler into many small passes
+that are easier for student to digest~\citep{Sarkar:2004fk}.
+
+This paper presents a proof of correctness in Agda for four nanopasses
+that translate a source language with let binding, integer arithmetic,
+conditional expressions and Booleans into an x86-flavored register
+transfer language. These four passes correspond to the
+following three passes in \emph{Essentials of Compilation}:
+\begin{enumerate}
+\item Remove Complex Operands
+\item Explicate Control
+\item Select Instructions
+\end{enumerate}
+The process of proving of correctness of the Explicate Control pass
+made it clear to me that it has two distinct reponsibilities that
+deserve to be separated:
+\begin{itemize}
+\item Replacing \texttt{let} expressions with assignment statements.
+\item Translating Boolean expressions and conditionals into
+  \texttt{goto}-based control-flow.
+\end{itemize}
+(The RTLgen pass of CompCert also handles the above two
+responsibilities.)  So in this paper, the Explicate Control pass is
+split into two passes, the first named Lift Locals and the second that
+keeps the name Explicate Control.
+\begin{enumerate}
+\item Remove Complex Operands
+\item Lift Locals
+\item Explicate Control
+\item Select Instructions
+\end{enumerate}
+
+%% differs in that I also use a big-step semantics for the register
+%% transfer language (instead of small step), which streamlines some of
+%% the reasoning.
 

### 2025-06-19 paper.tex
check

@@
 
+\usepackage{mathpartir}
+\usepackage{listings}
+\usepackage{stmaryrd}
+
+\usepackage[utf8]{inputenc}
+%\usepackage[T1]{fontenc}
+
+%\usepackage[verbose]{newunicodechar}
+%% \newunicodechar{ℕ}{\ensuremath{\mathbb{N}~}}
+%% \newunicodechar{𝔹}{\ensuremath{\mathbb{B}~}}
+%\newunicodechar{ℤ}{\ensuremath{\mathbb{Z}~}}
+%\newunicodechar{→}{\ensuremath{~\to~}}
+%\newunicodechar{×}{\ensuremath{~\times~}}
+
+\newcommand{\LangIf}{$\Lang_{\mathsf{If}}$} %R2
+\newcommand{\RCO}{\mathit{mon}} % output of remove-complex-opera*
+\newcommand{\LangIfANF}{\ensuremath{\Lang^{\RCO}_{\mathsf{if}}}} %R2
+\newcommand{\CLang}{\mathcal{C}}
+\newcommand{\LangCIf}{$\CLang_{\mathsf{If}}$} %C1
+
 \newcommand{\TRUE}[0]{\mathsf{true}}
@@
 \newcommand{\CB}[3]{\mathcal{B} \llbracket #1 \rrbracket ( #2, #3 )}
+
+\lstset{%
+language=Python,
+basicstyle=\ttfamily\small,
+morekeywords={match,case,bool,int,let,begin,if,else,closure},
+deletekeywords={},
+escapechar=|,
+mathescape=true,
+columns=flexible,
+%moredelim=[is][\color{red}]{~}{~},
+showstringspaces=false,
+%inputencoding=utf8
+literate={→}{{\to}}1,
+}
 
@@
 \[
-(A \to C) \times (B \to C) \cong (A + B) \to C
+(A → C) × (B → C) ≅ (A + B) → C
 \]
@@
 
+TODO: roadmap
+
 %% differs in that I also use a big-step semantics for the register
@@
 %% the reasoning.
+
+\section{The Languages}
+
+In this section we describe how we formalize the syntax and semantics
+of the source language and four intermediate languages in Agda:
+\begin{enumerate}
+\item Integers, Booleans, \texttt{let}, and \texttt{if}
+\item Monadic Normal Form
+\item Imperative with Structured Control Flow
+\item A C-like language with \emph{goto}
+\item X86 assembly but with variables
+\end{enumerate}
+These languages and there semantics are quite standard. However, there
+are many other intermediate languages in the literature to choose
+from, so why these languages, and why in this particular order?  The
+answer is that they happen to represent the images of the four
+nanopasses.  So then the question becomes, why do we choose to order
+the nanopasses in this way? The main thing to consider is that some
+passes make later passes easier because they remove a language feature
+that would have been complicated to handle. For example, the Lift
+Locals pass removed \texttt{let}, and I found \texttt{let} difficult
+to deal with in the proof of correctness for Explicate Control
+(\texttt{let} is relatively easy to handle in the code for Explicate
+Control). The other thing to consider is that some compiler passes use
+certain language features in their generated code, so the passes that
+remove those language features better come later in the ordering.
+
+One of the challenges of formalizing semantics in a proof assistant is
+that one needs to choose a representation of variables. There are many
+options, and I've tried most of them. But just like a game of
+whac-a-mole, the savings acheived in one place are lost in others.  My
+current favorite are good old de Bruijn indices. The following type
+\lstinline{Id} is for identifiers.
+
+\begin{lstlisting}
+Id = $ℕ$
+\end{lstlisting}
+
+\subsection{The source language}
+
+The source language is a language with integers, Booleans,
+\texttt{Let} binding, and conditional expressions. We choose a couple
+representative primitive operators: \texttt{Sub} for integer
+subtraction and \texttt{Eq} for equality on integers and
+Booleans.\footnote{Adding more primitive operators is straightforward
+but uninteresting.}  The \texttt{Read} operation requests an integer
+input from the user of the program.
+
+\begin{lstlisting}
+data Exp : Set where
+  Num : $ℤ$ $→$ Exp
+  Bool : $𝔹$ $→$ Exp
+  Read : Exp
+  Sub : Exp $→$ Exp $→$ Exp
+  Eq : Exp $→$ Exp $→$ Exp
+  Var : Id $→$ Exp
+  Let : Exp $→$ Exp $→$ Exp
+  If : Exp $→$ Exp $→$ Exp $→$ Exp
+\end{lstlisting}
+
+Towards specifying the semantics of this language, the inputs to the
+program are an infinite stream of integers represented as a natural
+number and a function from the naturals to the integers.  The natural
+number is incremented each time the program reads an input.
+
+\begin{lstlisting}
+Inputs = $ℕ$ $×$ ($ℕ$ $→$ $ℤ$)
+\end{lstlisting}
+
+The following monad is used to thread the inputs through the
+interpreter and to account for partiality via \lstinline{Maybe}
+(e.g. looking up an undefined variable).
+
+\begin{lstlisting}
+Reader A = Inputs $→$ Maybe (A $×$ Inputs)
+\end{lstlisting}
+
+\noindent The result of an expression is an integer or Boolean, for which we
+define the \lstinline{Value} type.
+
+\begin{lstlisting}
+data Value : Set where
+  Int : $ℤ$ $→$ Value
+  Bool : $𝔹$ $→$ Value
+\end{lstlisting}
+
+\noindent The variables (de Bruijn indices) are interpreted using an
+environment that is represented as a list of values.
+
+\begin{lstlisting}
+Env A = List A
+\end{lstlisting}
+
+\noindent The semantics of expressions is given by the
+\lstinline{interp-exp} function, which maps an expression and
+environment to a value, in our \lstinline{Reader} monad.
+
+\begin{lstlisting}
+interp-exp : Exp $→$ Env Value $→$ Reader Value
+\end{lstlisting}
+
+\noindent The semantics of a whole program is given by the
+\lstinline{interp-LIf} function.
+
+\begin{lstlisting}
+interp-LIf : Exp $→$ Inputs $→$ Maybe Value
+\end{lstlisting}
+
+\subsection{Monadic normal form}
+
+The first intermediate language restricts the arguments of the
+primitives \lstinline{Sub} and \lstinline{Eq} to be \emph{atomic}
+expressions, which we define to include variables and literals. This
+restriction gets us closer to x86, where instruction arguments may
+only be immediates (literals) or registers.
+
+\begin{lstlisting}
+data Atm : Set where
+  Num : $ℤ$ $→$ Atm 
+  Bool : $𝔹$ $→$ Atm 
+  Var : Id $→$ Atm
+
+data Mon : Set where
+  Atom : Atm $→$ Mon
+  Read : Mon
+  Sub : Atm $→$ Atm $→$ Mon
+  Eq : Atm $→$ Atm $→$ Mon
+  Let : Mon $→$ Mon $→$ Mon
+  If : Mon $→$ Mon $→$ Mon $→$ Mon
+\end{lstlisting}
+
+The interpreter for atoms does not make use of the \lstinline{Reader}
+monad because atoms do not have side effects.
+
+\begin{lstlisting}
+interp-atm : Atm $→$ Env Value $→$ Maybe Value
+\end{lstlisting}
+
+The interpreter for \lstinline{Mon} is similar to the
+one for \lstinline{Exp}, except that it dispatches
+to \lstinline{interp-atm} for atoms. The semantics
+of a whole program is given by \lstinline{interp-LMonIf}.
+
+\begin{lstlisting}
+interp-mon : Mon $→$ Env Value $→$ Reader Value
+interp-LMonIf : Mon $→$ Inputs $→$ Maybe Value
+\end{lstlisting}
+
+\subsection{Structured imperative IR}
+
+The third language replaces the \lstinline{Let} bindings with
+assignment expressions but retains the structured control-flow and
+expression-oriented nature of the language. This language brings us
+closer to x86 because variables are represented by registers and stack
+locations, all of which are mutable. The abstract syntax for
+assignment is \lstinline{(Assign $x$ $e_1$ $e_2$)} where $e_1$ is the
+right-hand side and $e_2$ is an expression to be executed after the
+assignment (reminiscent of the body of a \lstinline{Let}).  The
+\lstinline{Program} AST node consists of the number of variables and
+an expression.
+
+\begin{lstlisting}
+data Imp-Exp : Set where
+  Atom : Atm $→$ Imp-Exp
+  Read : Imp-Exp
+  Sub : Atm $→$ Atm $→$ Imp-Exp
+  Eq : Atm $→$ Atm $→$ Imp-Exp
+  Assign : Id $→$ Imp-Exp $→$ Imp-Exp $→$ Imp-Exp
+  If : Imp-Exp $→$ Imp-Exp $→$ Imp-Exp $→$ Imp-Exp
+
+data Imp-Prog : Set where
+  Program : $ℕ$ $→$ Imp-Exp $→$ Imp-Prog
+\end{lstlisting}
+
+Regarding the semantics, the main change is that environments are now
+both an input and output. Also, in anticipation of the next language,
+we switch to representing the semantics as a relation, that is, a
+big-step semantics.
+
+\begin{lstlisting}
+StateImp = Inputs × Env Value
+data _$⊢$_$⇓$_$⊣$_ : StateImp $→$ Imp-Exp
+                     $→$ Value $→$ StateImp $→$ Set
+\end{lstlisting}
+
+\noindent Here is the rule for assignment:
+\[
+\inferrule{s , ρ ⊢ e_1 ⇓ v_1 ⊣ s′ , ρ′ \\
+           s′ , ρ′(x := v_1) ⊣ e_2 ⇓ s″ , ρ″}
+          {s , ρ ⊢ \mathtt{Assign}~ x~ e_1~ e_2 ⊣ s″ , ρ″}
+\]
+
+The semantics of whole programs is given by the following relation.
+
+\begin{lstlisting}
+interp-imp : Imp-Prog $→$ Inputs $→$ Value $→$ Set
+\end{lstlisting}
+
+\subsection{The C-like language}
+
+The fourth language is similar to the C programming language in that
+it supports \texttt{goto} and separates expressions from statements.
+However, to get us closer to x86, the condition of an \lstinline{if}
+statement is restricted to an equality test. This form of
+\lstinline{if} can be easily translated into a sequence of three x86
+instructions: \lstinline{cmpq}, \lstinline{je}, and \lstinline{jmp}.
+However, compiling from the previous IR to this one is non-trivial
+because the previous IR allowed arbitrarily nested expressions to
+appear in the condition of an \lstinline{if}, including other
+\lstinline{if}'s!
+
+\begin{lstlisting}
+data CExp : Set where
+  Atom : Atm $→$ CExp
+  Read : CExp
+  Sub : Atm $→$ Atm $→$ CExp
+  Eq : Atm $→$ Atm $→$ CExp
+
+data CStmt : Set where
+  Return : CExp $→$ CStmt
+  Assign : Id $→$ CExp $→$ CStmt $→$ CStmt
+  IfEq : Atm $→$ Atm $→$ Id $→$ Id $→$ CStmt
+  Goto : Id $→$ CStmt
+
+data CProg : Set where
+  Program : $ℕ$ $→$ Id $→$ List CStmt $→$ CProg
+\end{lstlisting}
+
+Note that the \lstinline{CStmt} type represents not just one statement
+but a block of statements that always end in either a
+\lstinline{Return} statement (that halts the program) or a jump to
+another block (\lstinline{Goto} or \lstinline{IfEq}).
+
+We choose a big-step semantics for C-like language because the
+semantics is no longer structurally recursive and furthermore,
+programs are no longer guaranteed to terminate.\footnote{The programs
+in the image of Explicate Control are guaranteed to terminate, but
+Agda doesn't know that.}  So encoding the semantics as a function in
+Agda would come with complications (such as using a gas parameter)
+that can be side-stepped by using a big-step semantics. Compared to
+the semantics for the imperative IR, this semantics refers to a list
+of blocks $B$ that can be the targets of gotos.
+
+\begin{lstlisting}
+data _,_$⊢ᶜ$_$⇓$_$⊣$_ : StateImp $→$ List CStmt $→$ CStmt
+                      $→$ Value $→$ StateImp $→$ Set
+\end{lstlisting}
+
+The rule for \lstinline{Goto} looks up the block for label $l$ and
+executes that block.
+\[
+\inferrule{s , B ⊢ᶜ B[l] ⇓ v ⊣ s′}
+          {s , B ⊢ᶜ \mathtt{Goto}~ l ⇓ v ⊣ s′}
+\]
+\noindent The rule for \lstinline{IfEq}, in the case when the
+arguments are equal, is as follows:
+\[
+\inferrule{\texttt{interp-CExp}(\mathtt{Eq}~a_1~a_2) = \mathsf{just}~\mathsf{true} \\
+           s , B ⊢ᶜ B[l_1] ⇓ v ⊣ s′}
+          {s , B ⊢ᶜ \mathtt{IfEq}~a_1~a_2~l_1~l_2 ⇓ v ⊣ s′}
+\]
+
+\noindent The semantics of whole programs is given by the following
+relation.
+
+\begin{lstlisting}
+interp-prog : CProg $→$ Inputs $→$ Value $→$ Set
+\end{lstlisting}
+
+\subsection{x86 with variables}
+
+The final intermediate language is a subset of x86 assembly language,
+except for the addition of variables. A complete compiler would
+include a few more passes to perform register allocation and deal with
+idiosyncrosies of x86, such as only allowing at most one instruction
+argument to access memory.
+
+\begin{lstlisting}
+data Arg : Set where
+  Num : $ℤ$ $→$ Arg
+  Bool : $𝔹$ $→$ Arg
+  Var : Id $→$ Arg
+  Reg : $ℕ$ $→$ Arg
+
+data Dest : Set where
+  Var : Id $→$ Dest
+  Reg : $ℕ$ $→$ Dest
+
+data Inst : Set where
+  MovQ : Arg $→$ Dest $→$ Inst
+  SubQ : Arg $→$ Dest $→$ Inst
+  CmpQ : Arg $→$ Arg $→$ Inst
+  Jmp : Id $→$ Inst
+  JmpEq : Id $→$ Inst
+  ReadInt : Inst
+
+Block : Set
+Block = List Inst
+
+data X86Var : Set where
+  Program : $ℕ$ $→$ Id $→$ List Block $→$ X86Var
+\end{lstlisting}
+
+The mutable state consists of the input, the registers, and the
+variables. However, we only use one register, \lstinline{rax}.
+
+%input + registers + variables
+\begin{lstlisting}
+StateX86 = Inputs × List Value × List Value
+\end{lstlisting}
+
+The meaning of each instruction is given by the following relation.
+The Boolean in this relation specifies whether the instruction
+falls-through to the next instruction or not (because it jumped).
+
+\begin{lstlisting}
+data _,_$⊢$_$⇓$_,_ : StateX86 $→$ List Block $→$ Inst
+                    $→$ StateX86 $→$ $𝔹$ $→$ Set
+\end{lstlisting}
+
+\noindent We take a few liberties to simplify the semantics with
+respect to x86 assembly language. For example, our \lstinline{CmpQ}
+instruction puts the result of the comparison into \lstinline{rax}
+instead of the special \lstinline{EFLAGS} register. Likewise,
+\lstinline{JmpEq} does not check the \lstinline{EFLAGS} register, but
+instead jumps if \lstinline{rax} contains \lstinline{true}.
+
+\noindent The meaning of a sequence of instructions is given the
+following big-step relation.
+
+\begin{lstlisting}
+data _,_$⊩$_$⇓$_,_ : StateX86 $→$ List Block $→$ List Inst
+                     $→$ StateX86 $→$ $𝔹$ $→$ Set 
+\end{lstlisting}
+
+This big-step treatment of the semantics of instruction sequences is
+unusual, so here are the three rules.
+
+\begin{gather*}
+\inferrule{}{s , B ⊩ [] ⇓ s , \mathsf{true}} \\[1ex]
+\inferrule{s , B ⊢ i ⇓ s′ , \mathsf{true} \\
+           s′ , B ⊩ is ⇓ s″ , b}
+          {s , B ⊩ i ∷ is ⇓ s′ , b}\\[1ex]
+\inferrule{s , B ⊢ i ⇓ s′ , \mathsf{false}}
+          {s , B ⊩ i ∷ is ⇓ s′ , b}
+\end{gather*}
+
+The meaning of a whole program is given by the following relation.
+\begin{lstlisting}
+interp-x86-var : X86Var → Inputs → Value → Set
+\end{lstlisting}
+
+We now proceed to discuss each compiler pass and its proof of
+correctness.
+
+\section{Pass: Remove Complex Operands}
+
+\section{Pass: Lift Locals}
+
+\section{Pass: Explicate Control}
+
+\section{Pass: Select Instructions}
+
+\section{The Compiler is Correct}
 

### 2025-06-19 paper.tex
lift locals

@@
 basicstyle=\ttfamily\small,
-morekeywords={match,case,bool,int,let,begin,if,else,closure},
-deletekeywords={},
-escapechar=|,
+keywordstyle=\ttfamily\small,
+%morekeywords={match,case,bool,int,let,begin,if,else,closure},
+%deletekeywords={locals},
+%escapechar=|,
 mathescape=true,
@@
 that would have been complicated to handle. For example, the Lift
-Locals pass removed \texttt{let}, and I found \texttt{let} difficult
+Locals pass removes \texttt{let}, and I found \texttt{let} difficult
 to deal with in the proof of correctness for Explicate Control
@@
 certain language features in their generated code, so the passes that
-remove those language features better come later in the ordering.
+remove those language features better come later in the ordering.  For
+example, the Closure Conversion pass (not discussed here) generates
+code with \texttt{let}, so Closure Conversion comes before Lift
+Locals.
 
@@
 
-\subsection{The source language}
+\subsection{The Source Language}
 
-The source language is a language with integers, Booleans,
-\texttt{Let} binding, and conditional expressions. We choose a couple
-representative primitive operators: \texttt{Sub} for integer
-subtraction and \texttt{Eq} for equality on integers and
-Booleans.\footnote{Adding more primitive operators is straightforward
-but uninteresting.}  The \texttt{Read} operation requests an integer
-input from the user of the program.
+The source language includes integers, Booleans, \texttt{Let} binding,
+and conditional expressions. We choose a couple representative
+primitive operators: \texttt{Sub} for integer subtraction and
+\texttt{Eq} for equality on integers and Booleans.\footnote{Adding
+more primitive operators is straightforward but uninteresting.}  The
+\texttt{Read} operation requests an integer input from the user of the
+program.\footnote{\texttt{Read} is necessary to prevent students from
+interpreting the programs instead of compiling them.}
 
@@
 
-\subsection{Monadic normal form}
+\subsection{Monadic Normal Form}
 
@@
 restriction gets us closer to x86, where instruction arguments may
-only be immediates (literals) or registers.
+only be immediates (literals), registers, or memory accesses.
 
@@
 
-\subsection{Structured imperative IR}
+\subsection{Structured Imperative IR}
 
@@
 
+We implement the Remove Complex Operands pass with the following
+\lstinline{rco} function in Agda. This pass is responsible for making
+sure that the arguments of primitive operations, such as
+\lstinline{Sub} and \lstinline{Eq}, are atomic expression (variables
+or literals).
+
+\begin{lstlisting}
+rco : Exp $→$ Mon
+\end{lstlisting}
+
+Most of the cases of \lstinline{rco} are trivial, but let us focus on
+the case for subtraction. We shall start with a strawman
+translation. We recursively compile the subexpressions $e_1$ and
+$e_2$. The results might be complex (not atomic), so we
+\lstinline{let} binding them both and replace them in the subtraction
+with variables (de Bruijn indices 0 and 1). Note that we must shift
+the variables in $e_2$ up by one because we inserted a new
+\lstinline{let} above $e_2$. (In general, \lstinline{shift-mon m c}
+increases by one each variable in $m$ that is greater or equal to the
+cutoff $c$.)
+
+\begin{lstlisting}
+rco (Sub e$₁$ e$₂$) =
+  Let (rco e$₁$)
+   (Let (shift-mon (rco e$₂$) 0)
+    (Sub (Var 1) (Var 0)))
+\end{lstlisting}
+
+However, the above translation inserts more \lstinline{let} bindings
+than are strictly necessary. If $e_1$ and/or $e_2$ compile to an atom,
+then there is no need to bind them to a \lstinline{let}.  The
+following is an alternative translation for \lstinline{Sub} that
+includes four different cases, depending on whether \lstinline{rco e$₁$}
+and \lstinline{rco e$₂$} are atomic.
+
+\begin{lstlisting}
+rco (Sub e$₁$ e$₂$)
+    with rco e$₁$ | rco e$₂$
+... | m$₁$ | m$₂$
+    with atomic? m$₁$ | atomic? m$₂$
+... | yes (atomic a$₁$) | yes (atomic a$₂$) =
+      Sub a$₁$ a$₂$
+... | no cmplx$₁$ | yes (atomic a$₂$) =
+      Let (rco e$₁$) (Sub (Var zero) (shift-atm a$₂$ 0))
+... | yes (atomic a$₁$) | no cmplx$₂$ =
+      Let (rco e$₂$) (Sub (shift-atm a$₁$ 0) (Var 0))
+... | no cmplx$₁$ | no cmplx$₂$ = 
+      Let m$₁$
+        (Let (shift-mon m$₂$ 0)
+          (Sub (Var 1) (Var 0)))
+\end{lstlisting}
+
+(It would be nice to collapse the four cases into one, but I have not
+yet figured out how to do that while accounting for the necessary
+shifting of de Bruijn indices.)
+
+The correctness theorem for Remove Complex Operands says that for any
+program $e$, interpreting $\mathtt{rco}\,e$ with input $s$ yields the
+same output as directly interpreting $e$ with the same input.
+
+\begin{lstlisting}
+rco-correct : $∀$ (e : Exp) (s : Inputs)
+  $→$ interp-LMonIf (rco e) s $≡$ interp-LIf e s 
+\end{lstlisting}
+
+The main lemma is that \lstinline{rco} followed by
+\lstinline{interp-mon} is equivalent to \lstinline{interp-exp}.
+
+\begin{lstlisting}
+rco-correct-exp : ∀ (e : Exp) (ρ : Env Value)
+  → interp-mon (rco e) ρ ≡ interp-exp e ρ
+\end{lstlisting}
+
+The proof is by structural induction on $e$, but it requires the
+following lemmas regarding the shifting of de Bruijn indices.
+The idea is that the result of looking up variable $x$ in
+environment ($ρ₁$ ++ $ρ₂$) produces the same result as
+shifting the variable by one with the cutoff set to the
+the length of $ρ₁$ and looking it up in the environment
+($ρ₁$ ++ v $∷$ $ρ₂$), for any value $v$.
+
+\begin{lstlisting}
+nth-++-shift-var : ($ρ₁$ $ρ₂$ : List A)(v : A)(x : Id)
+  $→$ nth ($ρ₁$ ++ v $∷$ $ρ₂$) (shift-var x (length $ρ₁$))
+    $≡$ nth ($ρ₁$ ++ $ρ₂$) x
+
+interp-shift-atm : (a : Atm)(v : Value)($ρ₁$ $ρ₂$ : Env Value)
+  $→$ interp-atm (shift-atm a (length $ρ₁$)) ($ρ₁$ ++ v $∷$ $ρ₂$) 
+    $≡$ interp-atm a ($ρ₁$ ++ $ρ₂$)
+
+interp-shift-mon : (m : Mon)(v : Value)($ρ₁$ $ρ₂$ : Env Value)
+  $→$ interp-mon (shift-mon m (length $ρ₁$)) ($ρ₁$ ++ (v $∷$ $ρ₂$))
+    $≡$ interp-mon m ($ρ₁$ ++ $ρ₂$)
+\end{lstlisting}
+
 \section{Pass: Lift Locals}
+
+The Lift Locals pass is responsible for replacing all of the
+\lstinline{let} expressions with assignment statements.  After this
+pass, variables are still represented by de Bruijn indices but they
+are all bound by the \lstinline{Program} AST node.
+
+The \lstinline{lift-locals} function delegates to the recursive
+function \lstinline{lift-locals-mon}, where all the work is done.
+That function returns a number $n$ that counts the number of
+\lstinline{let} bindings in the expression, and the translated
+expression.
+
+\begin{lstlisting}
+lift-locals : Mon → Imp-Prog
+lift-locals m
+    with lift-locals-mon m
+... | n , e = Program n e    
+\end{lstlisting}
+
+\noindent The case of \lstinline{lift-locals-mon} for \lstinline{Let}
+is the most interesting. One can imagine that the result of
+\lstinline{lift-locals-mon m$₁$} is a sequence of $i$ imaginary
+\lstinline{let}'s surrounding \lstinline{e$₁$}. Similarly for 
+\lstinline{lift-locals-mon m$₁$}.
+The translation result for this \lstinline{Let} should be
+a sequence of $i + j + 1$ imaginary
+\lstinline{let}'s, followed by an assignment
+of \lstinline{e$₁$} to the appropriate variable, and then the
+code for \lstinline{e$₂$}. However, we need to shift
+the variables in \lstinline{e$₁$} up by $j + 1$ because
+we are moving all of the $j$ \lstinline{let}'s for
+\lstinline{e$₂$} above \lstinline{e$₁$}, as well as the
+current \lstinline{let}. Thus, we need another function
+for shifting de Bruijn indices by an arbitrary amount.
+
+\begin{lstlisting}
+shifts-imp-exp : Imp-Exp $→$ $ℕ$ $→$ $ℕ$ $→$ Imp-Exp
+\end{lstlisting}
+
+The variables in \lstinline{e$₂$} also need to be shifted.  They are
+shifted up by $i$ because we are moving all of the $i$
+\lstinline{let}'s for \lstinline{e$₁$} above
+\lstinline{e$₂$}. Regarding the variable that serves as the left-hand
+side of the assignment, its de Bruijn index is $i + j$ because the
+original index for this \lstinline{Let} was $0$ but we have moved $i +
+j$ other \lstinline{let}s above this variable occurence.
+
+\begin{lstlisting}
+lift-locals-mon (Let m$₁$ m$₂$)
+    with lift-locals-mon m$₁$
+... | i , e$₁$
+    with lift-locals-mon m$₂$
+... | j , e$₂$ =
+    let e$′₁$ = shifts-imp-exp e$₁$ 0 (suc j) in
+    let e$′₂$ = shifts-imp-exp e$₂$ j i in
+    suc (i + j) , Assign (i + j) e$′₁$ e$′₂$
+\end{lstlisting}
+
+The case of \lstinline{lift-locals-mon} for \lstinline{If} is also
+complicated because there are three subexpressions that must all be
+shifted appropriately.
+
+The correctness theorem for Lift Locals says that if a program output
+is $v$ given input $s$, then Lift Locals produces a program that
+outputs the same value given the same input.
+
+\begin{lstlisting}
+lift-locals-correct : $∀$(m : Mon)(s : Inputs)(v : Value)
+  $→$ interp-LMonIf m s $≡$ just v
+  $→$ interp-imp (lift-locals m) s v
+\end{lstlisting}
+
+This theorem is a corollary of the following main lemma, which is
+proved by structural induction on the expression \lstinline{m}.
+Considerable care is needed in the statement of this lemma so that the
+induction hypotheses can be used. Recall that when interpreting an
+expression, the environment $ρ$ provides the meaning for the
+\emph{free variables} in the expression.  As the interpreter descends
+under \lstinline{let} bindings, it pushes values on the front of the
+environment. However, after Lift Locals, all of the variables are
+bound at the top, so the initial environment won't be empty, but
+instead will have entries for all of the variables (we arbitrarily
+initialize them to $0$). So in general, we can separate the
+environment into two halves, $ρ₁$ and $ρ₂$, where $ρ₂$ is for the free
+variables in \lstinline{m} whereas \lstinline{$ρ₁$ ++ $ρ₂$} is for the
+free variables in the result expressoin \lstinline{e} of
+\lstinline{lift-locals-mon m}. The length of $ρ₁$ corresponds to the
+number returned by \lstinline{lift-locals-mon m}. Also, note that
+because \lstinline{Mon} is a pure language, \lstinline{e} contains a
+single assignment to each variable, so $ρ₂$ is unchanged by the
+execution of \lstinline{e}. ($ρ₁$ is changed as each of its variables
+are assigned to for the first time.)
+
+\begin{lstlisting}
+lift-locals-mon-correct : $∀$(m : Mon)(e : Imp-Exp)
+    (s $s′$ : Inputs)(v : Value) ($ρ₁$ $ρ₂$ : Env Value)
+  $→$ interp-mon m $ρ₂$ s $≡$ just (v , $s′$)
+  $→$ lift-locals-mon m $≡$ (length $ρ₁$ , e)
+  $→$ $∃$[ $ρ′₁$ ] (s , $ρ₁$ ++ $ρ₂$) $⊢$ e $⇓$ v $⊣$ ($s′$ , $ρ′₁$ ++ $ρ₂$)
+      × length $ρ₁$ $≡$ length $ρ′₁$
+\end{lstlisting}
+
+The main lemma relies on more lemmas about shifting de Bruijn indices
+that generalize our previous ones to arbitrary amounts of shifting.
+
+\begin{lstlisting}
+$⇓$shifts : $∀${e}{v}{s $s′$}{$ρ₁$ $ρ′₁$ $ρ₂$ $ρ₃$ $ρ′₃$} 
+  $→$ (s , $ρ₁$++$ρ₃$) $⊢$ e $⇓$ v $⊣$ ($s′$ , $ρ′₁$++$ρ′₃$)
+  $→$ length $ρ′₁$ $≡$ length $ρ₁$
+  $→$ (s , $ρ₁$++$ρ₂$++$ρ₃$) $⊢$
+       shifts-imp-exp e (length $ρ₁$) (length $ρ₂$)
+       $⇓$ v $⊣$ ($s′$ , $ρ′₁$++$ρ₂$++$ρ′₃$)
+\end{lstlisting}
 

### 2025-06-19 paper.tex
check

@@
 $\mathcal{B}$ that takes two more arguments, the generated code for
-the two branches. Thus, $\mathcal{B}$ can analyze the expression $b$
-to decide how to generate code for the $\mathsf{if}$, or whether an
-$\mathsf{if}$ is even necessary.
+the two branches, that is, for the two continuations of $b$. Thus,
+$\mathcal{B}$ can analyze the expression $b$ to decide how to generate
+code for the $\mathsf{if}$, or whether an $\mathsf{if}$ is even
+necessary.
 \[

### 2025-06-20 paper.tex
tweaks

@@
 
+\usepackage{url}
 \usepackage{mathpartir}
@@
 case-of-case transformation is employed to optimize dispatching on
-algebraic datatypes \citep{PEYTONJONES19983}.
+algebraic datatypes~\citep{PEYTONJONES19983}.
 
 Olivier Danvy, in his paper \emph{A New One-Pass Transformation into
-Monadic Normal Form}~\citep{Danvy:2003fk}, presents a beautiful family
+Monadic Normal Form}~\citep{Danvy:2003fk}, presents an elegant family
 of structurally recursive functions that translate Boolean expressions
@@
 \subsection{Structured Imperative IR}
+\label{sec:imp}
 
@@
 
+CFG = List CStmt
+
 data CProg : Set where
-  Program : $ℕ$ $→$ Id $→$ List CStmt $→$ CProg
+  Program : $ℕ$ $→$ Id $→$ CFG $→$ CProg
 \end{lstlisting}
@@
 Note that the \lstinline{CStmt} type represents not just one statement
-but a block of statements that always end in either a
+but a sequence of statements that always end in either a
 \lstinline{Return} statement (that halts the program) or a jump to
-another block (\lstinline{Goto} or \lstinline{IfEq}).
+another \lstinline{CStmt} via \lstinline{Goto} or \lstinline{IfEq}.
+The control flow graph \lstinline{CFG} is a list of \lstinline{CStmt},
+and jump labels are indexes into that list.  So \lstinline{Program}
+consists of (1) the number of variables, (2) the index of the start
+node, (3) the control flow graph.
 
@@
 Agda would come with complications (such as using a gas parameter)
-that can be side-stepped by using a big-step semantics. Compared to
-the semantics for the imperative IR, this semantics refers to a list
-of blocks $B$ that can be the targets of gotos.
-
-\begin{lstlisting}
-data _,_$⊢ᶜ$_$⇓$_$⊣$_ : StateImp $→$ List CStmt $→$ CStmt
+that we side-step by using a big-step semantics. Compared to the
+semantics for the imperative IR, this semantics includes a parameter
+for the control flow graph.
+
+\begin{lstlisting}
+data _,_$⊢ᶜ$_$⇓$_$⊣$_ : StateImp $→$ CFG $→$ CStmt
                       $→$ Value $→$ StateImp $→$ Set
@@
 
-The rule for \lstinline{Goto} looks up the block for label $l$ and
-executes that block.
+The rule for \lstinline{Goto} looks up the CFG node for label $l$ and
+executes the sequence of statements in that node.
 \[
-\inferrule{s , B ⊢ᶜ B[l] ⇓ v ⊣ s′}
-          {s , B ⊢ᶜ \mathtt{Goto}~ l ⇓ v ⊣ s′}
+\inferrule{s , G ⊢ᶜ G[l] ⇓ v ⊣ s′}
+          {s , G ⊢ᶜ \mathtt{Goto}~ l ⇓ v ⊣ s′}
 \]
@@
 \inferrule{\texttt{interp-CExp}(\mathtt{Eq}~a_1~a_2) = \mathsf{just}~\mathsf{true} \\
-           s , B ⊢ᶜ B[l_1] ⇓ v ⊣ s′}
-          {s , B ⊢ᶜ \mathtt{IfEq}~a_1~a_2~l_1~l_2 ⇓ v ⊣ s′}
+           s , G ⊢ᶜ G[l_1] ⇓ v ⊣ s′}
+          {s , G ⊢ᶜ \mathtt{IfEq}~a_1~a_2~l_1~l_2 ⇓ v ⊣ s′}
 \]
@@
 shifting of de Bruijn indices.)
+
+\subsection{Remove Complex Operands is Correct}
 
@@
 shifted appropriately.
+
+\subsection{Lift Locals is Correct}
 
@@
 
+The Explicate Control pass is responsible for translating from the
+structured control-flow of an abstract syntax \emph{tree} into a
+language with a goto-based control flow \emph{graph}.\footnote{Our
+source language does not include loops, so the graph happens to be a
+directed acyclic graph.} In particular, Explicate Control translates
+\lstinline{If} expressions, whose conditions may be arbitrary
+expressions that include other \lstinline{If} expressions, into the
+\lstinline{IfEq} form that only branches based on an equality test. In
+so doing, Explicate Control brings together uses of the \lstinline{Eq}
+comparison operator and the \lstinline{If} expressions that depend on
+them, thereby avoiding the needless extra code that would be generated
+by a naive approach.
+
+We implement the Explicate Control pass with four functions in Agda,
+the \lstinline{explicate} function translates a whole program from the
+imperative IR to the C-like language, but it delegates all the work to
+the other three functions. These three functions are organized
+according to the context in which they are called. Reviewing the
+grammar of \lstinline{Imp-Exp} (section~\ref{sec:imp}), there are
+three contexts in which an \lstinline{Imp-Exp} may appear:
+\begin{enumerate}
+\item Tail position: the result of the expression is the output of the program,
+\item Assignment: the result of the expression is assigned to a variable, and
+\item Predicate: the (Boolean) result of the expression determines which
+    branch to take in a conditional.
+\end{enumerate}
+
+The functions for Assignment and Predicate have extra parameters for
+their continuations, one for Assignment and two for Predicate. The
+continuations are represented simply as the generated code in the
+C-like IL. The function for Tail position of course needs no
+continuation parameter.
+
+All three functions will be building a control flow graph, which needs
+to be threaded into and out of these functions, so they use the
+following a \lstinline{Build} monad:
+
+\begin{lstlisting}
+Build A = CFG $→$ A × CFG
+retB : {A : Set} $→$ A $→$ Build A
+_thenB_ : {A B : Set} $→$ Build A $→$ (A $→$ Build B)
+           $→$ Build B
+add-node : CStmt $→$ Build Id
+\end{lstlisting}
+
+\noindent Here are the signatures of the four functions that make up
+the Explicate Control pass. Next we discuss select cases of these
+functions.
+
+\begin{lstlisting}
+explicate : Imp-Prog $→$ CProg
+explicate-tail : Imp-Exp $→$ Build CStmt
+explicate-assign : Id $→$ Imp-Exp $→$ CStmt $→$ Build CStmt
+explicate-pred : Imp-Exp $→$ CStmt $→$ CStmt $→$ Build CStmt
+\end{lstlisting}
+
+\paragraph{Explicate Tail}
+
+Starting with \lstinline{explicate-tail}, in the case of an atomic
+expression or primitive operator, we generate a return statement.
+Here is the case for subtraction.
+
+\begin{lstlisting}
+explicate-tail (Sub a$₁$ a$₂$) = retB (Return (Sub a$₁$ a$₂$))
+\end{lstlisting}
+
+\noindent Moving on to the assignment statement, we see a pattern of
+compiling the program in reverse. We first compile the continuation
+\lstinline{e$₂$}, which is in tail position, producing the code \lstinline{c$₂$}.
+We then compile \lstinline{e$₁$}, which is in assignment position,
+passing \lstinline{c$₂$} as its continuation.
+
+\begin{lstlisting}
+explicate-tail (Assign x e$₁$ e$₂$) =
+    (explicate-tail e$₂$) thenB
+    $λ$ c$₂$ $→$ explicate-assign x e$₁$ c$₂$
+\end{lstlisting}
+
+\noindent Regarding a conditional expression, we first compile the two
+branches, \lstinline{e$₂$} and \lstinline{e$₃$}, which are both in tail
+position, and then compile the condition \lstinline{e$₁$}, which is in
+predicate position, so we pass \lstinline{c$₂$} and \lstinline{c$₃$} in
+the continuation parameters.
+
+\begin{lstlisting}
+explicate-tail (If e$₁$ e$₂$ e$₃$) =
+  (explicate-tail e$₂$) thenB
+  $λ$ c$₂$ $→$ (explicate-tail e$₃$) thenB
+  $λ$ c$₃$ $→$ explicate-pred e$₁$ c$₂$ c$₃$
+\end{lstlisting}
+
+\paragraph{Explicate Assignments}
+
+Turning to \lstinline{explicate-assign}, in each base case (for
+atomics and primitive operators), we simply generate an assignment
+statement with continuation \lstinline{c}.
+
+\begin{lstlisting}
+explicate-assign x (Atom a) c =
+  retB (Assign x (Atom a) c)
+\end{lstlisting}
+
+The case for conditional expressions is interesting: we do not
+directly generate an assignment statement. The two branches,
+\lstinline{e$₂$} and \lstinline{e$₃$}, are in assignment position,
+so we recursively compile them with \lstinline{explicate-assign}.
+They both have the same continuation \lstinline{c} and the first rule
+of compilation is that you cannot duplicate code.\footnote{GHC
+violated this rule a decade ago:
+\url{https://stackoverflow.com/questions/35815503/what-ghc-optimization-is-responsible-for-duplicating-case-expressions}.}
+Thus, we add \lstinline{c} to the control-flow graph and
+instead use a \lstinline{Goto} to its label as the continuation
+for \lstinline{e$₂$} and \lstinline{e$₃$}.
+
+\begin{lstlisting}
+explicate-assign y (If e$₁$ e$₂$ e$₃$) c =
+   add-node c thenB
+   $λ$ l $→$ explicate-assign y e$₂$ (Goto l) thenB
+   $λ$ c$₂$ $→$ explicate-assign y e$₃$ (Goto l) thenB
+   $λ$ c$₃$ $→$ explicate-pred e$₁$ c$₂$ c$₃$
+\end{lstlisting}
+
+\paragraph{Explicate Predicates}
+
+The base cases for \lstinline{explicate-pred} are interesting.  If the
+expression is a literal \lstinline{true} or \lstinline{false}, then we
+return the continuation for the \lstinline{thn} or \lstinline{els}
+branch, respectively. (This may case \lstinline{thn} or
+\lstinline{els} to become dead code, which we discuss later.)
+
+\begin{lstlisting}
+explicate-pred (Atom (Bool true)) thn els = retB thn
+explicate-pred (Atom (Bool false)) thn els = retB els
+\end{lstlisting}
+
+\noindent In the case of a variable, we compare it to \lstinline{true}
+and branch to either the \lstinline{thn} or \lstinline{els}
+continuations. However, the syntax of \lstinline{IfEq} requires labels
+to jump to, so we add both continuations to the CFG.
+
+\begin{lstlisting}
+explicate-pred (Atom (Var x)) thn els =
+  add-node thn thenB
+  $λ$ l$₁$ $→$ add-node els thenB
+  $λ$ l$₂$ $→$ retB (IfEq (Var x) (Bool true) l$₁$ l$₂$)
+\end{lstlisting}
+
+\noindent The case for an equality comparison delivers on one of the
+goals of Explicate Control, which is to combine comparison expressions
+with branching whenever possible.
+
+\begin{lstlisting}
+explicate-pred (Eq a$₁$ a$₂$) thn els =
+  add-node thn thenB
+  $λ$ l$₁$ $→$ add-node els thenB
+  $λ$ l$₂$ $→$ retB (IfEq a$₁$ a$₂$ l$₁$ l$₂$)
+\end{lstlisting}
+
+For the finale of the Explicate Control pass, we discuss the
+translation of conditionals in \lstinline{explicate-pred}.  The two
+branches \lstinline{e$₂$} and \lstinline{e$₃$} are in predicate
+position, so we compile them with recursive calls to
+\lstinline{explicate-pred} to obtain \lstinline{c$₂$} and
+\lstinline{c$₃$}. However, they share the same two continuations,
+\lstinline{thn} and \lstinline{els}, so we add both of them to the CFG
+to avoid code duplication. Finally, we compile \lstinline{e$₁$} with
+\lstinline{explicate-pred}, passing \lstinline{c$₂$} and
+\lstinline{c$₃$} as the continuations.
+
+\begin{lstlisting}
+explicate-pred (If e$₁$ e$₂$ e$₃$) thn els =
+   add-node thn thenB
+   $λ$ l$₁$ $→$ add-node els thenB
+   $λ$ l$₂$ $→$ explicate-pred e$₂$ (Goto l$₁$) (Goto l$₂$) thenB
+   $λ$ c$₂$ $→$ (explicate-pred e$₃$ (Goto l$₁$) (Goto l$₂$)) thenB
+   $λ$ c$₃$ $→$ explicate-pred e$₁$ c$₂$ c$₃$
+\end{lstlisting}
+
+\noindent The above handling of \lstinline{If} happens to perform the
+case-of-case optimization~\citep{PEYTONJONES19983}.
+
+\paragraph{Discussion}
+
+Throughout the Explicate Control pass we added code to the
+control-flow graph whenever we needed to avoid code
+duplication. However, some further care is needed to avoid adding (1)
+trivial nodes that only contain a \lstinline{Goto} and (2) nodes that
+are never the target of a jump (i.e., dead code). The
+\lstinline{add-node} function can easily handle (1) by inspecting its
+argument and if it is a \lstinline{Goto}, just return its label
+instead of adding a new node to the CFG.  Avoiding dead code (2) is
+more challenging, either requiring a post-processing of the CFG or use
+lazy evaluation to delay the addition of nodes to the CFG to the point
+where they are definitely needed~\citep{Siek:2023tr}.
+
+\subsection{Explicate Control is Correct}
+
+The correctness theorem for Explicate Control says that if a program
+output is $v$ given input $s$, then Explicate Control produces a
+program that outputs the same value given the same input.
+
+\begin{lstlisting}
+explicate-correct : $∀$ {p}{s}{v}
+  $→$ interp-imp p s v
+  $→$ interp-prog (explicate p) s v
+\end{lstlisting}
+
+The correctness theorem delegates to three lemmas, one for each
+\lstinline{explicate-*} function.
+
+\paragraph{Explicate Tail is Correct}
+
+The statement of the lemma for \lstinline{explicate-tail} is the most
+straightforward. If expression \lstinline{e} compiles to
+\lstinline{c}, and if \lstinline{e} evaluates to \lstinline{v}, then
+\lstinline{c} also evaluates to \lstinline{v}:
+\begin{lstlisting}
+explicate-tail-correct :
+  $∀${e}{$ρ$ $ρ'$}{s s'}{G G$′$}{c}{v}
+  $→$ (s , $ρ$) $⊢$ e $⇓$ v $⊣$ (s' , $ρ'$)
+  $→$ explicate-tail e G $≡$ (c , G$′$)
+  $→$ (s , $ρ$) , G$′$ $⊢ᶜ$ c $⇓$ v $⊣$ (s' , $ρ'$)
+\end{lstlisting}
+\noindent The cases in the proof of \lstinline{explicate-tail-correct}
+for assignment and conditional expressions rely on the correctness
+lemmas for \lstinline{explicate-assign} and
+\lstinline{explicate-pred}.
+
+The case for conditional expressions also requires a lemma about the
+monotonicity of the control flow graph.  The induction hypothesis for
+\lstinline{e$₂$} says that the generated \lstinline{c$₂$} evaluates
+using the CFG that was produced by the compilation of
+\lstinline{e$₂$}, but to satisfy the induction hypothesis
+for \lstinline{e$₁$} we need \lstinline{c$₂$} evaluated using the CFG
+that was produced by the compilation of \lstinline{e$₃$}.
+We define the following notation for
+expression that one CFG is a prefix of another CFG:
+\begin{lstlisting}
+G$₁$ $↝$ G$₂$ = $∃$[ G ] G$₁$ ++ G $≡$ G$₂$
+\end{lstlisting}
+\noindent We prove the following lemma which states that the ouput CFG of
+\lstinline{explicate-tail} is an extension of its input CFG.
+\begin{lstlisting}
+explicate-tail-CFG : $∀${e}{G$₁$ G$₂$}{c}
+  $→$ explicate-tail e G$₁$ $≡$ (c , G$₂$)
+  $→$ G$₁$ $↝$ G$₂$
+\end{lstlisting}
+
+\paragraph{Explicate Assign is Correct}
+
+The correctness lemma for \lstinline{explicate-assign} adds another
+premise which states that the continuation \lstinline{c$₂$} evaluates
+in an environment that was updated by the assignment.
+
+\begin{lstlisting}
+explicate-assign-correct :
+  $∀${x}{e$₁$}{c$₂$}{$ρ$ $ρ′$ $ρ″$}{s s$′$ s$″$}{G G$′$}{c$₁$}{n v}
+  $→$ (s , $ρ$) $⊢$ e$₁$ $⇓$ n $⊣$ (s$′$ , $ρ′$)
+  $→$ (s$′$ , update $ρ′$ x n) , G $⊢ᶜ$ c$₂$ $⇓$ v $⊣$ (s$″$ , $ρ″$)
+  $→$ explicate-assign x e$₁$ c$₂$ G $≡$ (c$₁$ , G$′$)
+  $→$ (s , $ρ$) , G$′$ $⊢ᶜ$ c$₁$ $⇓$ v $⊣$ (s$″$ , $ρ″$)
+\end{lstlisting}
+
+\noindent The case for conditional expressions needs a couple more
+lemmas regarding control flow graphs. Similar to
+\lstinline{explicate-tail}, we need a lemma which states that the
+ouput CFG of \lstinline{explicate-assign} is an extension of its input
+CFG.
+
+\begin{lstlisting}
+explicate-assign-CFG : ∀{x}{m}{t t'}{G₁ G₂}
+  → explicate-assign x m t G₁ ≡ (t' , G₂)
+  → G₁ ↝ G₂
+\end{lstlisting}
+
+\paragraph{Explicate Predicate is Correct}
+
+...
+
+\begin{lstlisting}
+explicate-pred-correct :
+  $∀$ {e$₁$}{c$₁$ c$₂$ c}{b}{v}{s s$′$ s$″$}{$ρ$ $ρ′$ $ρ″$}{G$′$ G$″$ G$‴$}
+  $→$ (s , $ρ$) $⊢$ e$₁$ $⇓$ Bool b $⊣$ (s$′$ , $ρ′$)
+  $→$ ((s$′$ , $ρ′$) , G$′$ $⊢ᶜ$ (if b then c$₂$ else c$₃$) $⇓$ v $⊣$ (s$″$ , $ρ″$))
+  $→$ G$′$ $↝$ G$″$
+  $→$ explicate-pred e$₁$ c$₂$ c$₃$ G$″$ $≡$ (c$₁$ , G$‴$)
+  $→$ (s , $ρ$) , G$‴$ $⊢ᶜ$ c$₁$ $⇓$ v $⊣$ (s$″$ , $ρ″$)
+\end{lstlisting}
+
 \section{Pass: Select Instructions}

### 2025-06-20 paper.tex
finished section on explicate

@@
 
-The case for conditional expressions also requires a lemma about the
+The case for conditional expressions also two lemmas about the
 monotonicity of the control flow graph.  The induction hypothesis for
@@
 \end{lstlisting}
-\noindent We prove the following lemma which states that the ouput CFG of
-\lstinline{explicate-tail} is an extension of its input CFG.
+\noindent We prove that the ouput CFG of \lstinline{explicate-tail} is
+an extension of its input CFG.
 \begin{lstlisting}
@@
   $→$ G$₁$ $↝$ G$₂$
+\end{lstlisting}
+\noindent And we prove that the big-step semantics is monotonic with
+respect to the $↝$ relation on control flow graphs.
+\begin{lstlisting}
+eval-CFG : $∀${c}{$ρ$ $ρ′$}{B$₁$ B$₂$}{s s$′$}{v}
+  → B$₁$ $↝$ B$₂$
+  → (s , $ρ$) , B$₁$ $⊢ᶜ$ c $⇓$ v $⊣$ (s$′$ , $ρ′$)
+  → (s , $ρ$) , B$₂$ $⊢ᶜ$ c $⇓$ v $⊣$ (s$′$ , $ρ′$)
 \end{lstlisting}
@@
 
-\noindent The case for conditional expressions needs a couple more
-lemmas regarding control flow graphs. Similar to
-\lstinline{explicate-tail}, we need a lemma which states that the
-ouput CFG of \lstinline{explicate-assign} is an extension of its input
-CFG.
-
-\begin{lstlisting}
-explicate-assign-CFG : ∀{x}{m}{t t'}{G₁ G₂}
-  → explicate-assign x m t G₁ ≡ (t' , G₂)
-  → G₁ ↝ G₂
+\noindent The case for conditional expressions needs another lemma
+regarding control flow graphs. Similar to \lstinline{explicate-tail},
+we need a lemma which states that the ouput CFG of
+\lstinline{explicate-assign} is an extension of its input CFG.
+
+\begin{lstlisting}
+explicate-assign-CFG : $∀${x}{m}{c c'}{G$₁$ G$₂$}
+  → explicate-assign x m c G$₁$ $≡$ (c' , G$₂$)
+  → G$₁$ $↝$ G$₂$
 \end{lstlisting}
@@
 
-...
+The following is the statement of correctness for
+\lstinline{explicate-pred-correct}.
 
@@
 explicate-pred-correct :
-  $∀$ {e$₁$}{c$₁$ c$₂$ c}{b}{v}{s s$′$ s$″$}{$ρ$ $ρ′$ $ρ″$}{G$′$ G$″$ G$‴$}
+  $∀$ {e$₁$}{c$₁$ c$₂$ c}{b}{v}{s s$′$ s$″$}{$ρ$ $ρ′$ $ρ″$}{G G$′$}
   $→$ (s , $ρ$) $⊢$ e$₁$ $⇓$ Bool b $⊣$ (s$′$ , $ρ′$)
-  $→$ ((s$′$ , $ρ′$) , G$′$ $⊢ᶜ$ (if b then c$₂$ else c$₃$) $⇓$ v $⊣$ (s$″$ , $ρ″$))
-  $→$ G$′$ $↝$ G$″$
-  $→$ explicate-pred e$₁$ c$₂$ c$₃$ G$″$ $≡$ (c$₁$ , G$‴$)
-  $→$ (s , $ρ$) , G$‴$ $⊢ᶜ$ c$₁$ $⇓$ v $⊣$ (s$″$ , $ρ″$)
+  $→$ ((s$′$,$ρ′$) , G $⊢ᶜ$ if b then c$₂$ else c$₃$ $⇓$ v $⊣$ (s$″$,$ρ″$))
+  $→$ explicate-pred e$₁$ c$₂$ c$₃$ G $≡$ (c$₁$ , G$′$)
+  $→$ (s , $ρ$) , G$′$ $⊢ᶜ$ c$₁$ $⇓$ v $⊣$ (s$″$ , $ρ″$)
+\end{lstlisting}
+
+\noindent This statement of correctness is interesting because it
+requires the premise that either continuation \lstinline{c$₂$} or
+\lstinline{c$₃$} evaluates to \lstinline{v}, depending on whether the
+expression \lstinline{e$₁$} evaluates to \lstinline{true} or
+\lstinline{false}. This explains the use of Agda's
+\lstinline{if-then-else} in the second premise of
+\lstinline{explicate-pred-correct}.\footnote{The formulation of this
+lemma follows that of CompCert's \lstinline{transl\_condexpr\_prop}.}
+
+The proof of \lstinline{explicate-pred-correct} uses many of the
+monotonicity lemmas that we have already proved, plus one more.
+
+\begin{lstlisting}
+explicate-pred-CFG : $∀${e}{c$₁$ c$₂$ c}{G$₁$ G$₂$}
+  $→$ explicate-pred e c$₁$ c$₂$ G$₁$ $≡$ (c , G$₂$)
+  $→$ G$₁$ $↝$ G$₂$
 \end{lstlisting}

### 2025-06-20 paper.tex
minor edit

@@
 
-%% \author{Lars Th{\o}rv{\"a}ld}
-%% \affiliation{%
-%%   \institution{The Th{\o}rv{\"a}ld Group}
-%%   \city{Hekla}
-%%   \country{Iceland}}
-%% \email{larst@affiliation.org}
-
-%% \author{Valerie B\'eranger}
-%% \affiliation{%
-%%   \institution{Inria Paris-Rocquencourt}
-%%   \city{Rocquencourt}
-%%   \country{France}
-%% }
-
-%% \author{Aparna Patel}
-%% \affiliation{%
-%%  \institution{Rajiv Gandhi University}
-%%  \city{Doimukh}
-%%  \state{Arunachal Pradesh}
-%%  \country{India}}
-
-%% \author{Huifen Chan}
-%% \affiliation{%
-%%   \institution{Tsinghua University}
-%%   \city{Haidian Qu}
-%%   \state{Beijing Shi}
-%%   \country{China}}
-
-%% \author{Charles Palmer}
-%% \affiliation{%
-%%   \institution{Palmer Research Laboratories}
-%%   \city{San Antonio}
-%%   \state{Texas}
-%%   \country{USA}}
-%% \email{cpalmer@prl.com}
-
-%% \author{John Smith}
-%% \affiliation{%
-%%   \institution{The Th{\o}rv{\"a}ld Group}
-%%   \city{Hekla}
-%%   \country{Iceland}}
-%% \email{jsmith@affiliation.org}
-
-%% \author{Julius P. Kumquat}
-%% \affiliation{%
-%%   \institution{The Kumquat Consortium}
-%%   \city{New York}
-%%   \country{USA}}
-%% \email{jpkumquat@consortium.net}
-
 %%
@@
 
-%%
-%% The abstract is a short summary of the work to be presented in the
-%% article.
 \begin{abstract}
@@
 Intel's x86, the naive approach to compiling boolean expressions and
-conditionals produces embarassingly inefficient code. Similar issues
+conditionals produces embarrassingly inefficient code. Similar issues
 arise in the compilation of languages such as Haskell, where the
@@
 the Chez Scheme compiler~\citep{Dybvig:2006aa}.  One of those
-techniques is called destintation-driven code
-generation~\cite{Dybvig:1990aa}, which adds an extra parameter to the
+techniques is called destination-driven code
+generation~\citep{Dybvig:1990aa}, which adds an extra parameter to the
 compilation function to indicate how the context will use the result
@@
 languages, each of which adds a language feature to the previous
-langauge. For the language with Boolean expressions and conditionals,
+language. For the language with Boolean expressions and conditionals,
 the Explicate Control pass of the compiler uses the one-function per
@@
 The process of proving of correctness of the Explicate Control pass
-made it clear to me that it has two distinct reponsibilities that
+made it clear to me that it has two distinct responsibilities that
 deserve to be separated:
@@
 
-TODO: roadmap
-
-%% differs in that I also use a big-step semantics for the register
-%% transfer language (instead of small step), which streamlines some of
-%% the reasoning.
+Section~\ref{sec:langs} of the paper introduces the intermediate
+languages. Section~\ref{sec:compiler} describes the compiler,
+detailing each pass and the proof of correctness.
 
 \section{The Languages}
+\label{sec:langs}
 
@@
 
-One of the challenges of formalizing semantics in a proof assistant is
-that one needs to choose a representation of variables. There are many
-options, and I've tried most of them. But just like a game of
-whac-a-mole, the savings acheived in one place are lost in others.  My
-current favorite are good old de Bruijn indices. The following type
-\lstinline{Id} is for identifiers.
+One of the perennial challenges of formalizing semantics in a proof
+assistant is that one needs to choose a representation of variables.
+%% There are many
+%% options, and I've tried most of them. But just like a game of
+%% whac-a-mole, the savings achieved in one place are lost in others.
+Of the many options, my current favorite is good old de Bruijn
+indices~\citep{Bruijn:1972kx}.  Thus, identifiers \lstinline{Id} are
+represented as natural numbers.
 
@@
 \end{lstlisting}
+
+(This intermediate language does not correspond to the monadic normal
+form in Olivier's paper. It is the language of section
+\ref{sec:c-lang} that corresponds to the target of Olivier's one-pass
+transformation.)
 
@@
 \subsection{The C-like language}
+\label{sec:c-lang}
 
@@
 include a few more passes to perform register allocation and deal with
-idiosyncrosies of x86, such as only allowing at most one instruction
+idiosyncrasies of x86, such as only allowing at most one instruction
 argument to access memory.
@@
 
-We now proceed to discuss each compiler pass and its proof of
-correctness.
-
-\section{Pass: Remove Complex Operands}
+We now proceed to discuss the compiler and its proof of correctness.
+
+\section{The Compiler}
+\label{sec:compiler}
+
+The compiler is simply the composition of the four passes,
+which we detail in the next four subsections.
+
+\begin{lstlisting}
+compile e = (select-inst ∘ explicate ∘ lift-locals ∘ rco) e
+\end{lstlisting}
+
+\subsection{Pass: Remove Complex Operands}
 
@@
 Most of the cases of \lstinline{rco} are trivial, but let us focus on
-the case for subtraction. We shall start with a strawman
+the case for subtraction. We shall start with a straw-man
 translation. We recursively compile the subexpressions $e_1$ and
@@
 
-\subsection{Remove Complex Operands is Correct}
+\subsubsection{Remove Complex Operands is Correct}
 
@@
 
-\section{Pass: Lift Locals}
+\subsection{Pass: Lift Locals}
 
@@
 original index for this \lstinline{Let} was $0$ but we have moved $i +
-j$ other \lstinline{let}s above this variable occurence.
+j$ other \lstinline{let}s above this variable occurrence.
 
@@
 
-\subsection{Lift Locals is Correct}
+\subsubsection{Lift Locals is Correct}
 
@@
 variables in \lstinline{m} whereas \lstinline{$ρ₁$ ++ $ρ₂$} is for the
-free variables in the result expressoin \lstinline{e} of
+free variables in the result expression \lstinline{e} of
 \lstinline{lift-locals-mon m}. The length of $ρ₁$ corresponds to the
@@
 
-\section{Pass: Explicate Control}
+\subsection{Pass: Explicate Control}
 
@@
 according to the context in which they are called. Reviewing the
-grammar of \lstinline{Imp-Exp} (section~\ref{sec:imp}), there are
+grammar of \lstinline{Imp-Exp} (subsection~\ref{sec:imp}), there are
 three contexts in which an \lstinline{Imp-Exp} may appear:
@@
 
-\subsection{Explicate Control is Correct}
+\subsubsection{Explicate Control is Correct}
 
@@
 \end{lstlisting}
-\noindent We prove that the ouput CFG of \lstinline{explicate-tail} is
+\noindent We prove that the output CFG of \lstinline{explicate-tail} is
 an extension of its input CFG.
@@
 regarding control flow graphs. Similar to \lstinline{explicate-tail},
-we need a lemma which states that the ouput CFG of
+we need a lemma which states that the output CFG of
 \lstinline{explicate-assign} is an extension of its input CFG.
@@
 
-\section{Pass: Select Instructions}
-
-\section{The Compiler is Correct}
-
-%%
+\subsection{Pass: Select Instructions}
+
+The responsibility of the Select Instructions pass is to convert the
+assignment statements and primitive operations of the C-like language
+into x86 instructions. The minor challenge is that many x86
+instructions take just two arguments and have an update-in-place
+semantics.
+
+The function \lstinline{select-inst} translation a whole program in
+the C-like language to x86.
+\begin{lstlisting}
+select-inst (Program n l B) =
+  Program n l (map select-stmt B)
+\end{lstlisting}
+
+\noindent The \lstinline{select-stmt} function maps each statement to
+one or more instructions. The return statement is compiled like an
+assignment to the \lstinline{rax} register. (The calling convention
+for x86 says that the return value of a function is passed in
+\lstinline{rax}.)
+
+\begin{lstlisting}
+select-stmt (Return e) = select-assign e (Reg rax)
+\end{lstlisting}
+
+\noindent For the assignment statement we delegate to the
+\lstinline{select-assign} function, which analyzes the expression
+\lstinline{e} to determine which instruction to generate.
+We describe \lstinline{select-assign} shortly.
+
+\begin{lstlisting}
+select-stmt (Assign x e c) =
+  select-assign e (Var x) ++ select-stmt c
+\end{lstlisting}
+
+\noindent The \lstinline{IfEq} of the C-like language
+is translated to a \lstinline{CmpQ} instruction followed
+by \lstinline{JmpEq} and an unconditional \lstinline{Jmp}.
+
+\begin{lstlisting}
+select-stmt (IfEq a₁ a₂ thn els) =
+  CmpQ (to-arg a₁) (to-arg a₂) ∷
+  JmpEq thn ∷
+  Jmp els ∷ []
+\end{lstlisting}
+
+\noindent Finally, we translate \lstinline{Goto} into a \lstinline{Jmp}.
+\begin{lstlisting}
+select-stmt (Goto l) = Jmp l ∷ [] 
+\end{lstlisting}
+
+Next we look at some representative cases of
+\lstinline{select-assign}. The first parameter is the right-hand side
+of the assignment and the second parameter is the left-hand side, that
+is, the destination. For subtraction, we use the \lstinline{rax}
+register as scratch space. We first move argument \lstinline{a$₁$}
+into \lstinline{rax}, then subtract \lstinline{a$₂$} from
+\lstinline{rax}, and finally move the result into \lstinline{dest}.
+
+\begin{lstlisting}
+select-assign (Sub a$₁$ a$₂$) dest =
+  MovQ (to-arg a$₁$) (Reg rax) $∷$
+  SubQ (to-arg a$₂$) (Reg rax) $∷$
+  MovQ (Reg rax) dest $∷$ []
+\end{lstlisting}
+
+\noindent Of course there are special cases that require fewer
+instructions. For example, if the first argument of the subtraction is
+identical to the destination, then we can simply generate one
+\lstinline{SubQ} instruction.
+
+\begin{lstlisting}
+select-assign (Sub (Var x) a$₂$) (Var x) = 
+  SubQ (to-arg a$₂$) (Var x) $∷$ []
+\end{lstlisting}
+
+\noindent For the sake of keeping the proof short, we did not include
+any of the special cases.
+
+We compile the assignment of an equality comparison into a
+\lstinline{CmpQ} and a \lstinline{MovQ}. The \lstinline{CmpQ} always
+puts its result into \lstinline{rax}, so the \lstinline{MovQ} is
+needed to place the result into the destination.\footnote{The real
+\lstinline{cmpq} instruction is slightly stranger, involving
+the \lstinline{EFLAGS} register.}
+
+\begin{lstlisting}
+select-assign (Eq a$₁$ a$₂$) dest =
+  CmpQ (to-arg a$₁$) (to-arg a$₂$) $∷$
+  MovQ (Reg rax) dest $∷$ []
+\end{lstlisting}
+
+\subsubsection{Select Instructions is Correct}
+
+The correctness theorem for Select Instructions says that if a program
+output is $v$ given input $s$, then Explicate Control produces a
+program that outputs the same value given the same input.
+
+\begin{lstlisting}
+select-inst-correct : $∀${p}{s}{v}
+  → interp-prog p s v
+  → interp-x86-var (select-inst p) s v
+\end{lstlisting}
+
+The correctness theorem delegates to the correctness lemma for
+\lstinline{select-stmt}, which says that if statement \lstinline{c}
+evaluates to \lstinline{v}, then the instructions generated for
+\lstinline{c} put the same value \lstinline{v} into \lstinline{rax}.
+The list of x86 instruction blocks is obtain by mapping
+\lstinline{select-stmt} over the control-flow graph \lstinline{G}.
+
+\begin{lstlisting}
+select-stmt-correct : $∀$ (c : CStmt) ($ρ$ $ρ′$ : Env Value)
+  (s s$′$ : Inputs)(regs : List Value)(v : Value)(G : CFG)
+  $→$ (s , $ρ$) , G $⊢ᶜ$ c $⇓$ v $⊣$ (s$′$ , $ρ′$)
+  $→$ 0 < length regs
+  $→$ $∃$[ regs$′$ ] $∃$[ b ] (s , regs , $ρ$) , map select-stmt G
+         $⊩$ select-stmt c $⇓$ (s$′$ , regs$′$ , $ρ′$) , b
+    × nth regs$′$ rax $≡$ just v
+\end{lstlisting}
+
+When the statement is an assignment, the proof uses the following
+lemma. It states that if expression \lstinline{e} evaluates to
+\lstinline{v}, then the instructions generated for \lstinline{e} put
+\lstinline{v} into the specified destination. The predicate
+\lstinline{wrote} specifies the update to either the appropriate
+variable or register depending on the kind of destination.
+
+\begin{lstlisting}
+select-assign-correct : $∀$ (e : CExp) ($ρ$ : Env Value)
+   (s s$′$ : Inputs) (dest : Dest) (regs : List Value)
+   (v : Value) (B : List Block) 
+  $→$ interp-CExp e $ρ$ s $≡$ just (v , s$′$)
+  $→$ 0 < length regs
+  $→$ DestOK dest 1
+  $→$ $∃$[ st$′$ ] (s , regs , $ρ$) , B $⊩$
+            select-assign e dest $⇓$ st$′$ , true
+      × wrote dest v (s$′$ , regs , $ρ$) st$′$
+\end{lstlisting}
+
+\noindent The premise \lstinline{0 < length regs} ensures that we at
+least have one register (\lstinline{rax}). The premise
+\lstinline{DestOK dest 1} ensures that the destination is either a
+variable or the register (\lstinline{rax}).
+
+\subsection{The Compiler is Correct}
+
+The correctness theorem for the compiler composes the four correctness
+lemmas of the previous subsections.
+
+\begin{lstlisting}
+compile-correct : (e : Exp) (s : Inputs) (v : Value)
+  $→$ interp-LIf e s $≡$ just v
+  $→$ interp-x86-var (compile e) s v
+compile-correct e s v ie =
+   let i-rco = trans (rco-correct e s) ie in
+   let i-lift = lift-locals-correct (rco e) s v i-rco in
+   let i-exp = explicate-correct i-lift in
+   select-inst-correct i-exp
+\end{lstlisting}
+
+\section{Conclusion}
+
+This paper takes a step in adapting the literature on compiler
+correctness~\citep{Leroy:2006fe,Kumar:2014aa} to a university course
+based on the incremental and nanopass approach of \emph{Essentials of
+Compilation}~\citep{Siek:2023tr,Siek:2023ue}. We prove the correctness
+of a compiler for a source language with let binding, integer
+arithmetic, conditional expressions and Booleans, targeting an
+x86-flavored register transfer language. The process of developing
+mechanizing the proof in Agda encouraged a further split in the
+nanopasses, separating Explicate Control into two passes, one that
+lifts \lstinline{let} bindings and the second that compiles structured
+control-flow into \lstinline{goto}. In designing these compiler passes
+and proofs, it is enjoyable to trace the history of ideas going back
+to \citet{Aho:1986qf}, \citet{Dybvig:1990aa}, and Olivier Danvy's
+one-pass transformation into monadic normal form~\citep{Danvy:2003fk}.
+
 %% The next two lines define the bibliography style to be used, and
@@
 %% End of file `sample-sigplan.tex'.
+
+% LocalWords:  sigplan tex docstrip dtx bibtex TeXCount TC envir Siek
+% LocalWords:  displaymath Nanopasses authornote authornotemark Tobin
+% LocalWords:  rv Trovato al Agda nanopasses Olivier Danvy's ccs desc
+% LocalWords:  nanopass Ichiro Aho qf gotos datatypes PEYTONJONES fk
+% LocalWords:  Danvy arity Dybvig's Dybvig aa Chez Olivier's CompCert
+% LocalWords:  fe Coq RTLgen CminorSel CompCert's CakeML stackLang tr
+% LocalWords:  labLang Kumar ue Sarkar whac de Bruijn kx Eq Exp Num
+% LocalWords:  Bool Var Int Env interp exp LIf immediates Atm atm mon
+% LocalWords:  LMonIf IR AST Prog StateImp cmpq je jmp CExp CStmt CFG
+% LocalWords:  IfEq CProg prog Arg Reg Dest Inst MovQ SubQ CmpQ JmpEq
+% LocalWords:  ReadInt rax StateX EFLAGS var inst rco cmplx suc IL ie
+% LocalWords:  retB thenB pred GHC thn eval Agda's transl condexpr
+% LocalWords:  stmt arg dest regs DestOK trans

### 2025-08-11 paper.tex
check
added a running example

@@
 %deletekeywords={locals},
-%escapechar=|,
+escapechar=|,
 mathescape=true,
@@
 nanopass approach, breaking up the compiler into many small passes
-that are easier for student to digest~\citep{Sarkar:2004fk}.
+that are easier for student to digest~\citep{Sarkar:2004fk}.  There
+are many other pedogogical treatments of
+compilation~\citep{Appel:2003fk,Sumii:2005aa,Aho:2006wb,Cooper:2011aa,Grune:2012aa};
+what makes the \emph{Essentials of Compilation} unique is that it is
+the first textbook to apply the incremental approach.  However,
+\emph{Essentials of Compilation} does not include material about the
+formal verification of compilers; hence my present interest.
 
@@
 conditional expressions and Booleans into an x86-flavored register
-transfer language. These four passes correspond to the
+transfer language. The Agda code is available in the supplementary
+material. The four compiler passes presented here correspond to the
 following three passes in \emph{Essentials of Compilation}:
@@
 \end{enumerate}
+In compilers that do not take the nanopass approach, unrelated
+responsibilities are sometimes combined into the same pass.  For
+example, \citet{Sumii:2005aa} combines the first three of the above
+passes into a single pass.
 
@@
 \end{enumerate}
-These languages and there semantics are quite standard. However, there
+These languages and their semantics are quite standard. However, there
 are many other intermediate languages in the literature to choose
-from, so why these languages, and why in this particular order?  The
-answer is that they happen to represent the images of the four
-nanopasses.  So then the question becomes, why do we choose to order
-the nanopasses in this way? The main thing to consider is that some
-passes make later passes easier because they remove a language feature
-that would have been complicated to handle. For example, the Lift
-Locals pass removes \texttt{let}, and I found \texttt{let} difficult
-to deal with in the proof of correctness for Explicate Control
-(\texttt{let} is relatively easy to handle in the code for Explicate
-Control). The other thing to consider is that some compiler passes use
-certain language features in their generated code, so the passes that
-remove those language features better come later in the ordering.  For
-example, the Closure Conversion pass (not discussed here) generates
-code with \texttt{let}, so Closure Conversion comes before Lift
-Locals.
+from~\citep{Flanagan:1993cg,Hatcliff:1994vn,Kelsey:1995uq,Birkedal:1996aa},
+so why these languages, and why in this particular order?  The answer
+is that they happen to represent the images of the four nanopasses.
+So then the next question is: why do we order the nanopasses in this
+way? The main thing to consider is that some passes make later passes
+easier because they remove a language feature that would have been
+complicated to handle. For example, the Lift Locals pass removes
+\texttt{let}, and I found \texttt{let} difficult to deal with in the
+proof of correctness for Explicate Control (\texttt{let} is relatively
+easy to handle in the code for Explicate Control). The other thing to
+consider is that some compiler passes use certain language features in
+their generated code, so the passes that remove those language
+features better come later in the ordering.  For example, the Closure
+Conversion pass (not discussed here) generates code with \texttt{let},
+so Closure Conversion comes before Lift Locals. Finally, many design
+choices in \emph{Essentials of Compilation} have been inherited from
+\citet{Dybvig:2010aa}
 
@@
 assistant is that one needs to choose a representation of variables.
-%% There are many
-%% options, and I've tried most of them. But just like a game of
-%% whac-a-mole, the savings achieved in one place are lost in others.
 Of the many options, my current favorite is good old de Bruijn
@@
 
-\begin{lstlisting}
-compile e = (select-inst ∘ explicate ∘ lift-locals ∘ rco) e
-\end{lstlisting}
+\begin{lstlisting}[basicstyle=\ttfamily\footnotesize]
+compile e = (select-inst $∘$ explicate $∘$ lift-locals $∘$ rco) e
+\end{lstlisting}
+
+To illustrate each pass, we use the source program in
+figure~\ref{fig:running-example-source} as a running example,
+presented in a concrete syntax with variables as names instead of de
+Bruijn indices to enhance readability.  Given the inputs
+\lstinline{5}, \lstinline{10}, then \lstinline{0}, the result of this
+program is \lstinline{3}.
+
+\begin{figure}[tp]
+  \hrule
+\begin{lstlisting}
+    let x = read in
+    let y = read in
+    if (if x == 0
+        then true
+        else read == 1)
+    then y - 1
+    else (y - 2) - x
+\end{lstlisting}
+  \hrule
+\caption{An example source program.}
+\label{fig:running-example-source}
+\end{figure}
 
@@
 
+The Remove Complex Operands pass is responsible for making sure that
+the arguments of primitive operations, such as \lstinline{Sub} and
+\lstinline{Eq}, are atomic expression (variables or literals). It does
+so by insertion \lstinline{let} bindings for new temporary variables.
+In the running example (figure~\ref{fig:running-example-monadic}), the
+subexpressions \lstinline{read} and \lstinline{y - 2} are both
+arguments of primitive operations, so the Remove Complex Operands pass
+binds them to the temporary variables \lstinline{t1} and
+\lstinline{t2}, respectively.
+
+\begin{figure}[tp]
+  \hrule
+\begin{lstlisting}
+    let x = read in
+    let y = read in
+    if (if x == 0
+        then true
+        else (let t1 = read in t1 == 1))
+    then (y - 1)
+    else (let t2 = y - 2 in t2 - x)
+\end{lstlisting}
+  \hrule
+\caption{Example after Remove Complex Operands, in
+  monadic normal form.}
+\label{fig:running-example-monadic}
+\end{figure}
+
 We implement the Remove Complex Operands pass with the following
-\lstinline{rco} function in Agda. This pass is responsible for making
-sure that the arguments of primitive operations, such as
-\lstinline{Sub} and \lstinline{Eq}, are atomic expression (variables
-or literals).
-
+\lstinline{rco} function in Agda.
 \begin{lstlisting}
@@
 $e_2$. The results might be complex (not atomic), so we
-\lstinline{let} binding them both and replace them in the subtraction
+\lstinline{let}-bind them both and replace them in the subtraction
 with variables (de Bruijn indices 0 and 1). Note that we must shift
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[basicstyle=\ttfamily\footnotesize]
 nth-++-shift-var : ($ρ₁$ $ρ₂$ : List A)(v : A)(x : Id)
@@
 
-The Lift Locals pass is responsible for replacing all of the
-\lstinline{let} expressions with assignment statements.  After this
-pass, variables are still represented by de Bruijn indices but they
-are all bound by the \lstinline{Program} AST node.
-
+The Lift Locals pass is responsible for replacing \lstinline{let}
+expressions with assignment statements. In the output language,
+variables are declared at the top of the program.  Here is the result
+of Lift Locals on the running example. The output of Lift Locals on
+the running example is shown in
+figure~\ref{fig:running-example-structured-imperative}.
+
+\begin{figure}[tp]
+  \hrule
+\begin{lstlisting}
+    var x, y, t1, t2;
+    x := read;
+    y := read;
+    if (if x == 0
+        then true
+        else (t1 := read; t1 == 1)
+    then (y - 1)
+    else (t2 := y - 2; t2 - x)
+\end{lstlisting}
+  \hrule
+  \caption{Example after Lift Locals, in an imperative IR.}
+  \label{fig:running-example-structured-imperative}
+\end{figure}
 The \lstinline{lift-locals} function delegates to the recursive
@@
 
+The output of Explicate Control on the running example is shown in
+figure~\ref{fig:running-example-goto}. Note that there are two
+\lstinline{goto} statements that target \lstinline{block_2}, so indeed
+the control-flow graph of this program is a directed acyclic graph
+(DAG) and not a tree. The DAG representation enables sharing so the
+pass can avoid the duplication of code. We discuss the trivial blocks
+4 and 5 at the end of this section.
+
+\begin{figure}[tp]
+  \hrule
+\begin{lstlisting}
+    var x, y, t1, t2;
+    block_0:
+       x := read;
+       y := read;
+       if (x == 0) goto block_1; else goto block_3;
+    block_1:
+        goto block_2;
+    block_2:
+        return y - 1;
+    block_3:
+       t1 := read;
+       if (t1 == 1) goto block_4; else goto block_5;
+    block_4:
+        goto block_2;
+    block_5:
+        goto block_6;
+    block_6:
+        t2 := y - 2;
+        return t2 - x;
+\end{lstlisting}
+  \hrule
+  \caption{Example after Explicate Control, with control flow
+    made explicit with goto.}
+\label{fig:running-example-goto}
+\end{figure}
+
 We implement the Explicate Control pass with four functions in Agda,
@@
 
-\noindent Here are the signatures of the four functions that make up
-the Explicate Control pass. Next we discuss select cases of these
-functions.
-
-\begin{lstlisting}
+\noindent The following four functions make up the Explicate Control
+pass. Next we discuss select cases of these functions.
+
+\begin{lstlisting}[basicstyle=\ttfamily\footnotesize]
 explicate : Imp-Prog $→$ CProg
@@
 \lstinline{add-node} function can easily handle (1) by inspecting its
-argument and if it is a \lstinline{Goto}, just return its label
-instead of adding a new node to the CFG.  Avoiding dead code (2) is
-more challenging, either requiring a post-processing of the CFG or use
-lazy evaluation to delay the addition of nodes to the CFG to the point
-where they are definitely needed~\citep{Siek:2023tr}.
+argument and, if it is a \lstinline{Goto}, return its label instead of
+adding a new node to the CFG. This optimization is known as
+compressing corridor transitions~\citep{Danvy:2009aa}.  Avoiding dead
+code (2) is more challenging, either post-process the CFG or use lazy
+evaluation to delay the addition of nodes to the CFG to the time when
+they are definitely needed~\citep{Siek:2023tr}.
 
@@
 eval-CFG : $∀${c}{$ρ$ $ρ′$}{B$₁$ B$₂$}{s s$′$}{v}
-  → B$₁$ $↝$ B$₂$
-  → (s , $ρ$) , B$₁$ $⊢ᶜ$ c $⇓$ v $⊣$ (s$′$ , $ρ′$)
-  → (s , $ρ$) , B$₂$ $⊢ᶜ$ c $⇓$ v $⊣$ (s$′$ , $ρ′$)
+  $→$ B$₁$ $↝$ B$₂$
+  $→$ (s , $ρ$) , B$₁$ $⊢ᶜ$ c $⇓$ v $⊣$ (s$′$ , $ρ′$)
+  $→$ (s , $ρ$) , B$₂$ $⊢ᶜ$ c $⇓$ v $⊣$ (s$′$ , $ρ′$)
 \end{lstlisting}
@@
 regarding control flow graphs. Similar to \lstinline{explicate-tail},
-we need a lemma which states that the output CFG of
-\lstinline{explicate-assign} is an extension of its input CFG.
+we need a lemma which states that the output CFG is an extension of
+its input CFG.
 
@@
 explicate-assign-CFG : $∀${x}{m}{c c'}{G$₁$ G$₂$}
-  → explicate-assign x m c G$₁$ $≡$ (c' , G$₂$)
-  → G$₁$ $↝$ G$₂$
+  $→$ explicate-assign x m c G$₁$ $≡$ (c' , G$₂$)
+  $→$ G$₁$ $↝$ G$₂$
 \end{lstlisting}
@@
 instructions take just two arguments and have an update-in-place
-semantics.
+semantics. The output of Select Instructions on the running example is
+shown in figure~\ref{fig:running-example-x86}.\footnote{The redundant
+\texttt{movq} instructions are removed by the Patch Instructions pass
+in \emph{Essentials of Compilation}, which comes after Select
+Instructions and Register Allocation.}
+
+\begin{figure}[tp]
+  \hrule
+\begin{lstlisting}[mathescape=false]
+    var x, y, t1, t2
+    block_0:
+       callq read_int
+       movq %rax, x
+       callq read_int
+       movq %rax, y
+       cmpq x, $0
+       je block_1
+       jmp block_3
+    block_1:
+        jmp block_2
+    block_2:
+        movq y, %rax
+        subq $1, %rax
+        movq %rax, %rax
+    block_3:
+        callq read_int
+        movq %rax, t1
+        cmpq t1, $1
+        je block_4
+        jmp block_5
+    block_4:
+        jmp block_2
+    block_5:
+        jmp block_6
+    block_6:
+        movq y, %rax
+        subq $2, %rax
+        movq %rax, t2
+        movq t2, %rax
+        subq x, %rax
+        movq %rax, %rax
+\end{lstlisting}
+  \hrule
+  \caption{Example after Select Instructions, in the x86 assembly language
+    augmented with variables.}
+\label{fig:running-example-x86}
+\end{figure}
 
@@
 \begin{lstlisting}
-select-stmt (IfEq a₁ a₂ thn els) =
-  CmpQ (to-arg a₁) (to-arg a₂) ∷
-  JmpEq thn ∷
-  Jmp els ∷ []
+select-stmt (IfEq a$_1$ a$_2$ thn els) =
+  CmpQ (to-arg a$_1$) (to-arg a$_2$) $∷$
+  JmpEq thn $∷$
+  Jmp els $∷$ []
 \end{lstlisting}
@@
 \begin{lstlisting}
-select-stmt (Goto l) = Jmp l ∷ [] 
+select-stmt (Goto l) = Jmp l $∷$ [] 
 \end{lstlisting}
@@
 select-inst-correct : $∀${p}{s}{v}
-  → interp-prog p s v
-  → interp-x86-var (select-inst p) s v
+  $→$ interp-prog p s v
+  $→$ interp-x86-var (select-inst p) s v
 \end{lstlisting}

### 2025-08-11 paper.tex
add citation to continuation based PE
improvements from reviewers

@@
 \end{align*}
-
-My first exposure to these ideas were in Kent Dybvig's compiler
+\noindent The treatment code as continuation goes back to earlier work
+by \citet{Lawall:1994aa} on continuation-based partial
+evaluation. Indeed, the above is a partial evaluation of Boolean
+expressions.
+
+My first exposure to these ideas was in Kent Dybvig's compiler
 course~\citep{Dybvig:2010aa} where he taught the techniques used in
@@
 The CompCert compiler \citep{Leroy:2006fe} (a compiler for C verified
-in Coq) uses the one-function per context approach of Danvy in the
+in Rocq) uses the one-function per context approach of Danvy in the
 RTLgen pass that translates from the CminorSel intermediate language
@@
 compiler, on the other hand, employs a naive translation of Boolean
-expressions and conditionals for the compilation of stackLang the
+expressions and conditionals for the compilation of the stackLang 
 intermediate language into labLang~\citep{Kumar:2014aa}.
@@
 transfer language. The Agda code is available in the supplementary
-material. The four compiler passes presented here correspond to the
+material. It does not depend on particular features of Agda and could
+have been carried out in Rocq, Isabelle, or Lean.\footnote{The choice
+of Agda for this paper is not an endorsement of Agda but is a
+byproduct of the author's relative expertise in Agda at the time of
+writing.}
+
+The four compiler passes presented here correspond to the
 following three passes in \emph{Essentials of Compilation}:
@@
 \end{enumerate}
-The process of proving of correctness of the Explicate Control pass
+The process of proving correctness of the Explicate Control pass
 made it clear to me that it has two distinct responsibilities that
@@
 The source language includes integers, Booleans, \texttt{Let} binding,
-and conditional expressions. We choose a couple representative
-primitive operators: \texttt{Sub} for integer subtraction and
-\texttt{Eq} for equality on integers and Booleans.\footnote{Adding
-more primitive operators is straightforward but uninteresting.}  The
-\texttt{Read} operation requests an integer input from the user of the
+and conditional expressions. We choose two representative primitive
+operators: \texttt{Sub} for integer subtraction and \texttt{Eq} for
+equality on integers and Booleans.\footnote{Adding more primitive
+operators is straightforward but uninteresting.}  The \texttt{Read}
+operation requests an integer input from the user of the
 program.\footnote{\texttt{Read} is necessary to prevent students from
@@
 \begin{lstlisting}
-interp-x86-var : X86Var → Inputs → Value → Set
+interp-x86-var : X86Var $→$ Inputs $→$ Value $→$ Set
 \end{lstlisting}
@@
 \begin{lstlisting}
-rco-correct-exp : ∀ (e : Exp) (ρ : Env Value)
-  → interp-mon (rco e) ρ ≡ interp-exp e ρ
+rco-correct-exp : $∀$ (e : Exp) ($ρ$ : Env Value)
+  $→$ interp-mon (rco e) $ρ$ $≡$ interp-exp e $ρ$
 \end{lstlisting}
@@
 \begin{lstlisting}
-lift-locals : Mon → Imp-Prog
+lift-locals : Mon $→$ Imp-Prog
 lift-locals m
@@
 $⇓$shifts : $∀${e}{v}{s $s′$}{$ρ₁$ $ρ′₁$ $ρ₂$ $ρ₃$ $ρ′₃$} 
-  $→$ (s , $ρ₁$++$ρ₃$) $⊢$ e $⇓$ v $⊣$ ($s′$ , $ρ′₁$++$ρ′₃$)
+  $→$ (s , $ρ₁$ ++ $ρ₃$) $⊢$ e $⇓$ v $⊣$ ($s′$ , $ρ′₁$ ++ $ρ′₃$)
   $→$ length $ρ′₁$ $≡$ length $ρ₁$
-  $→$ (s , $ρ₁$++$ρ₂$++$ρ₃$) $⊢$
-       shifts-imp-exp e (length $ρ₁$) (length $ρ₂$)
-       $⇓$ v $⊣$ ($s′$ , $ρ′₁$++$ρ₂$++$ρ′₃$)
+  $→$ (s , $ρ₁$ ++ $ρ₂$ ++ $ρ₃$)
+      $⊢$ shifts-imp-exp e (length $ρ₁$) (length $ρ₂$)
+      $⇓$ v $⊣$ ($s′$ , $ρ′₁$ ++ $ρ₂$ ++ $ρ′₃$)
 \end{lstlisting}
@@
 to be threaded into and out of these functions, so they use the
-following a \lstinline{Build} monad:
+following \lstinline{Build} monad:
 
@@
 \lstinline{c} put the same value \lstinline{v} into \lstinline{rax}.
-The list of x86 instruction blocks is obtain by mapping
+The list of x86 instruction blocks is obtained by mapping
 \lstinline{select-stmt} over the control-flow graph \lstinline{G}.
 
-\begin{lstlisting}
+\begin{lstlisting}[basicstyle=\ttfamily\footnotesize]
 select-stmt-correct : $∀$ (c : CStmt) ($ρ$ $ρ′$ : Env Value)
@@
 % LocalWords:  Danvy arity Dybvig's Dybvig aa Chez Olivier's CompCert
-% LocalWords:  fe Coq RTLgen CminorSel CompCert's CakeML stackLang tr
+% LocalWords:  fe Rocq RTLgen CminorSel CompCert's CakeML stackLang tr
 % LocalWords:  labLang Kumar ue Sarkar whac de Bruijn kx Eq Exp Num

### 2025-08-11 paper.tex
one more

@@
 \usepackage{stmaryrd}
+%\usepackage[none]{hyphenat}
 
@@
 Towards developing a course on verified compilation, I am interested
-in adapting verification techniques from CompCert to the compiler
+in adapting these verification techniques to the compiler
 presented in the \emph{Essentials of Compilation}

### 2025-08-12 paper.tex
edits

@@
 \end{align*}
-\noindent The treatment code as continuation goes back to earlier work
-by \citet{Lawall:1994aa} on continuation-based partial
-evaluation. Indeed, the above is a partial evaluation of Boolean
-expressions.
+\noindent This idea builds on earlier work by \citet{Lawall:1994aa} on
+continuation-based partial evaluation. Indeed, the above is a partial
+evaluation of Boolean expressions.
 
@@
 compiler, on the other hand, employs a naive translation of Boolean
-expressions and conditionals for the compilation of the stackLang 
-intermediate language into labLang~\citep{Kumar:2014aa}.
+expressions and conditionals for the compilation of stackLang into
+labLang~\citep{Kumar:2014aa}.
 
 Towards developing a course on verified compilation, I am interested
-in adapting these verification techniques to the compiler
-presented in the \emph{Essentials of Compilation}
-textbook~\citep{Siek:2023tr,Siek:2023ue}. That book takes an
-incremental approach to teaching compilers, that is, it guides the
-students through the creation of multiple compilers for a sequence of
-languages, each of which adds a language feature to the previous
-language. For the language with Boolean expressions and conditionals,
-the Explicate Control pass of the compiler uses the one-function per
-context approach of Danvy to translate into an x86-flavored register
-transfer language.  \emph{Essentials of Compilation} also employs the
-nanopass approach, breaking up the compiler into many small passes
-that are easier for student to digest~\citep{Sarkar:2004fk}.  There
-are many other pedogogical treatments of
-compilation~\citep{Appel:2003fk,Sumii:2005aa,Aho:2006wb,Cooper:2011aa,Grune:2012aa};
-what makes the \emph{Essentials of Compilation} unique is that it is
-the first textbook to apply the incremental approach.  However,
-\emph{Essentials of Compilation} does not include material about the
-formal verification of compilers; hence my present interest.
+in proving correctness of the compiler presented in the
+\emph{Essentials of Compilation}
+textbook~\citep{Siek:2023tr,Siek:2023ue}. While it does not include
+proofs of correctness for its compiler passes, the textbook is
+organized in a way that might facilitate the proofs of correctness.
+%
+\emph{Essentials of Compilation} takes an \emph{incremental} approach
+to teaching compilers, that is, it guides the students through the
+creation of multiple compilers for a sequence of languages, each of
+which adds a language feature to the previous language. For the
+language with Boolean expressions and conditionals, the Explicate
+Control pass of the compiler uses the one-function per context
+approach of Danvy to translate into an x86-flavored register transfer
+language.
+%
+In addition to being incremental, \emph{Essentials of Compilation}
+employs the \emph{nanopass} approach, breaking up the compiler into
+many small passes that are easier for students to
+digest~\citep{Sarkar:2004fk}.
+%
+There are many other pedogogical treatments of
+compilation~\citep{Appel:2003fk,Sumii:2005aa,Aho:2006wb,Cooper:2011aa,Grune:2012aa}
+but none of them take the incremental approach.
 
@@
 transfer language. The Agda code is available in the supplementary
-material. It does not depend on particular features of Agda and could
+material and at the following location.
+\begin{center}
+\url{https://github.com/jsiek/correct_compilers}
+\end{center}
+The Agda code does not depend on special features of Agda and could
 have been carried out in Rocq, Isabelle, or Lean.\footnote{The choice
@@
 languages. Section~\ref{sec:compiler} describes the compiler,
-detailing each pass and the proof of correctness.
+detailing each pass and the proof of
+correctness. Section~\ref{sec:conclusion} concludes the paper.
 
@@
 from~\citep{Flanagan:1993cg,Hatcliff:1994vn,Kelsey:1995uq,Birkedal:1996aa},
-so why these languages, and why in this particular order?  The answer
-is that they happen to represent the images of the four nanopasses.
-So then the next question is: why do we order the nanopasses in this
-way? The main thing to consider is that some passes make later passes
-easier because they remove a language feature that would have been
-complicated to handle. For example, the Lift Locals pass removes
-\texttt{let}, and I found \texttt{let} difficult to deal with in the
-proof of correctness for Explicate Control (\texttt{let} is relatively
-easy to handle in the code for Explicate Control). The other thing to
-consider is that some compiler passes use certain language features in
-their generated code, so the passes that remove those language
-features better come later in the ordering.  For example, the Closure
-Conversion pass (not discussed here) generates code with \texttt{let},
-so Closure Conversion comes before Lift Locals. Finally, many design
-choices in \emph{Essentials of Compilation} have been inherited from
-\citet{Dybvig:2010aa}
+so why these languages, and why this order?  The answer is that they
+happen to represent the images of the four nanopasses.  So then the
+next question is: why do we order the nanopasses this way? The main
+thing to consider is that some passes make later passes easier because
+they remove a language feature that would have been complicated to
+handle. For example, the Lift Locals pass removes \texttt{let}, and I
+found \texttt{let} difficult to deal with in the proof of correctness
+for Explicate Control. The other thing to consider is that some
+compiler passes use certain language features in their generated code,
+so the passes that remove those language features better come later in
+the ordering.  For example, the Closure Conversion pass (not discussed
+here) generates code with \texttt{let}, so Closure Conversion comes
+before Lift Locals. Finally, many design choices in \emph{Essentials
+of Compilation} have been inherited from \citet{Dybvig:2010aa}
 
@@
 represented as natural numbers.
-
 \begin{lstlisting}
@@
 \section{Conclusion}
+\label{sec:conclusion}
 

### 2025-08-12 paper.tex
progress

@@
 expression-oriented nature of the language. This language brings us
-closer to x86 because variables are represented by registers and stack
+closer to x86 where variables are replaced by registers and stack
 locations, all of which are mutable. The abstract syntax for
@@
 appear in the condition of an \lstinline{if}, including other
-\lstinline{if}'s!
+\lstinline{if}s!
 
@@
 
-We choose a big-step semantics for C-like language because the
+We choose a big-step semantics for the C-like language because the
 semantics is no longer structurally recursive and furthermore,
@@
 The final intermediate language is a subset of x86 assembly language,
-except for the addition of variables. A complete compiler would
-include a few more passes to perform register allocation and deal with
-idiosyncrasies of x86, such as only allowing at most one instruction
-argument to access memory.
+except for the addition of variables. The compiler in \emph{Essentials
+of Compilation} includes a few more passes to perform register
+allocation and deal with the idiosyncrasies of x86, such as only
+allowing at most one instruction argument to access memory.
 
@@
 The Boolean in this relation specifies whether the instruction
-falls-through to the next instruction or not (because it jumped).
+falls through to the next instruction or not (because it jumped).
 
@@
 
-\noindent The meaning of a sequence of instructions is given the
+The meaning of a sequence of instructions is given the
 following big-step relation.
@@
 \end{lstlisting}
-
-This big-step treatment of the semantics of instruction sequences is
-unusual, so here are the three rules.
-
+This big-step treatment of instruction sequences is unusual, so here
+are the three rules.
 \begin{gather*}
@@
 \inferrule{s , B ⊢ i ⇓ s′ , \mathsf{false}}
-          {s , B ⊩ i ∷ is ⇓ s′ , b}
+          {s , B ⊩ i ∷ is ⇓ s′ , \mathsf{false}}
 \end{gather*}
-
-The meaning of a whole program is given by the following relation.
+The meaning of a program is given by the following relation.
 \begin{lstlisting}
@@
 
-We now proceed to discuss the compiler and its proof of correctness.
+Next we discuss the compiler and its proof of correctness.
 
@@
 
-The compiler is simply the composition of the four passes,
-which we detail in the next four subsections.
+The compiler is the composition of the four passes, which we detail in
+the next four subsections.
 
@@
 \lstinline{Eq}, are atomic expression (variables or literals). It does
-so by insertion \lstinline{let} bindings for new temporary variables.
+so by inserting \lstinline{let} bindings for new temporary variables.
 In the running example (figure~\ref{fig:running-example-monadic}), the
-subexpressions \lstinline{read} and \lstinline{y - 2} are both
+subexpressions \lstinline{read} and \lstinline{(y - 2)} are both
 arguments of primitive operations, so the Remove Complex Operands pass
@@
   \hrule
-\caption{Example after Remove Complex Operands, in
-  monadic normal form.}
+\caption{\small Result of Remove Complex Operands on the example.}
 \label{fig:running-example-monadic}
@@
 rco (Sub e$₁$ e$₂$)
-    with rco e$₁$ | rco e$₂$
-... | m$₁$ | m$₂$
-    with atomic? m$₁$ | atomic? m$₂$
-... | yes (atomic a$₁$) | yes (atomic a$₂$) =
+    with rco e$₁$ $|$ rco e$₂$
+... $|$ m$₁$ $|$ m$₂$
+    with atomic? m$₁$ $|$ atomic? m$₂$
+... $|$ yes (atomic a$₁$) $|$ yes (atomic a$₂$) =
       Sub a$₁$ a$₂$
-... | no cmplx$₁$ | yes (atomic a$₂$) =
+... $|$ no cmplx$₁$ $|$ yes (atomic a$₂$) =
       Let (rco e$₁$) (Sub (Var zero) (shift-atm a$₂$ 0))
-... | yes (atomic a$₁$) | no cmplx$₂$ =
+... $|$ yes (atomic a$₁$) $|$ no cmplx$₂$ =
       Let (rco e$₂$) (Sub (shift-atm a$₁$ 0) (Var 0))
-... | no cmplx$₁$ | no cmplx$₂$ = 
+... $|$ no cmplx$₁$ $|$ no cmplx$₂$ = 
       Let m$₁$
@@
 expressions with assignment statements. In the output language,
-variables are declared at the top of the program.  Here is the result
-of Lift Locals on the running example. The output of Lift Locals on
-the running example is shown in
+variables are declared at the top of the program.  The result of Lift
+Locals on the running example is shown in
 figure~\ref{fig:running-example-structured-imperative}.
@@
   \hrule
-  \caption{Example after Lift Locals, in an imperative IR.}
+  \caption{Result of Lift Locals on the example.}
   \label{fig:running-example-structured-imperative}
@@
 function \lstinline{lift-locals-mon}, where all the work is done.
-That function returns a number $n$ that counts the number of
-\lstinline{let} bindings in the expression, and the translated
-expression.
+That function returns (1) a number $n$ that counts the number of
+\lstinline{let} bindings that were in the expression and (2) the
+translation of the expression.
 
@@
     with lift-locals-mon m
-... | n , e = Program n e    
-\end{lstlisting}
-
-\noindent The case of \lstinline{lift-locals-mon} for \lstinline{Let}
-is the most interesting. One can imagine that the result of
-\lstinline{lift-locals-mon m$₁$} is a sequence of $i$ imaginary
+... $|$ n , e = Program n e    
+\end{lstlisting}
+
+\noindent The case of \lstinline{lift-locals-mon} for \lstinline{(Let m$_1$ m$_2$)}
+is the most interesting. Think of the result of
+\lstinline{lift-locals-mon m$₁$} as a sequence of $i$ imaginary
 \lstinline{let}'s surrounding \lstinline{e$₁$}. Similarly for
-\lstinline{lift-locals-mon m$₁$}.
-The translation result for this \lstinline{Let} should be
-a sequence of $i + j + 1$ imaginary
-\lstinline{let}'s, followed by an assignment
-of \lstinline{e$₁$} to the appropriate variable, and then the
-code for \lstinline{e$₂$}. However, we need to shift
-the variables in \lstinline{e$₁$} up by $j + 1$ because
-we are moving all of the $j$ \lstinline{let}'s for
-\lstinline{e$₂$} above \lstinline{e$₁$}, as well as the
-current \lstinline{let}. Thus, we need another function
-for shifting de Bruijn indices by an arbitrary amount.
+\lstinline{lift-locals-mon m$₁$}.  The translation result for
+\lstinline{(Let m$_1$ m$_2$)} should be a sequence of $i + j + 1$ imaginary
+\lstinline{let} bindings followed by an assignment of \lstinline{e$₁$} to
+the appropriate variable, and then the code for
+\lstinline{e$₂$}. However, we need to shift the variables in
+\lstinline{e$₁$} up by $j + 1$ because we are moving all of the $j$
+\lstinline{let} bindings for \lstinline{e$₂$} above \lstinline{e$₁$}, as well
+as the current \lstinline{let}. Thus, we need another function for
+shifting de Bruijn indices by an arbitrary amount.
 
@@
 shifted up by $i$ because we are moving all of the $i$
-\lstinline{let}'s for \lstinline{e$₁$} above
+\lstinline{let} bindings for \lstinline{e$₁$} above
 \lstinline{e$₂$}. Regarding the variable that serves as the left-hand
@@
     with lift-locals-mon m$₁$
-... | i , e$₁$
+... $|$ i , e$₁$
     with lift-locals-mon m$₂$
-... | j , e$₂$ =
+... $|$ j , e$₂$ =
     let e$′₁$ = shifts-imp-exp e$₁$ 0 (suc j) in
@@
 execution of \lstinline{e}. ($ρ₁$ is changed as each of its variables
-are assigned to for the first time.)
+are assigned for the first time.)
 
@@
 
-The main lemma relies on more lemmas about shifting de Bruijn indices
-that generalize our previous ones to arbitrary amounts of shifting.
+The main lemma relies on another lemma about shifting de Bruijn
+indices that generalizes our previous one to arbitrary amounts of
+shifting.
 
@@
   \hrule
-  \caption{Example after Explicate Control, with control flow
-    made explicit with goto.}
+  \caption{Example after Explicate Control, where control flow
+    is made explicit with goto.}
 \label{fig:running-example-goto}
@@
 
-All three functions will be building a control flow graph, which needs
-to be threaded into and out of these functions, so they use the
-following \lstinline{Build} monad:
+All three functions build a control flow graph, which needs to be
+threaded into and out of these functions, so they use the following
+\lstinline{Build} monad:
 
@@
 \noindent The following four functions make up the Explicate Control
-pass. Next we discuss select cases of these functions.
+pass. Next we discuss selected cases of these functions.
 
@@
 Thus, we add \lstinline{c} to the control-flow graph and
-instead use a \lstinline{Goto} to its label as the continuation
-for \lstinline{e$₂$} and \lstinline{e$₃$}.
+instead use a \lstinline{Goto} to the label for \lstinline{c}
+as the continuation for \lstinline{e$₂$} and \lstinline{e$₃$}.
 
@@
 For the finale of the Explicate Control pass, we discuss the
-translation of conditionals in \lstinline{explicate-pred}.  The two
+translation of the conditional \lstinline{(If e$₁$ e$₂$ e$₃$)}
+in \lstinline{explicate-pred}.  The two
 branches \lstinline{e$₂$} and \lstinline{e$₃$} are in predicate
@@
 \lstinline{thn} and \lstinline{els}, so we add both of them to the CFG
-to avoid code duplication. Finally, we compile \lstinline{e$₁$} with
+to avoid code duplication. We compile \lstinline{e$₁$} with
 \lstinline{explicate-pred}, passing \lstinline{c$₂$} and

### 2025-08-12 paper.tex
proof reading

@@
 straightforward. If expression \lstinline{e} compiles to
-\lstinline{c}, and if \lstinline{e} evaluates to \lstinline{v}, then
+\lstinline{c} and \lstinline{e} evaluates to \lstinline{v}, then
 \lstinline{c} also evaluates to \lstinline{v}:
@@
 
-The case for conditional expressions also two lemmas about the
+The case for conditional expressions requires two lemmas about the
 monotonicity of the control flow graph.  The induction hypothesis for
@@
 We define the following notation for
-expression that one CFG is a prefix of another CFG:
+saying that one CFG is a prefix of another CFG:
 \begin{lstlisting}
@@
 \end{lstlisting}
-\noindent And we prove that the big-step semantics is monotonic with
+\noindent We prove that the big-step semantics is monotonic with
 respect to the $↝$ relation on control flow graphs.
@@
   $→$ (s , $ρ$) $⊢$ e$₁$ $⇓$ Bool b $⊣$ (s$′$ , $ρ′$)
-  $→$ ((s$′$,$ρ′$) , G $⊢ᶜ$ if b then c$₂$ else c$₃$ $⇓$ v $⊣$ (s$″$,$ρ″$))
+  $→$ ((s$′$,$ρ′$) , G $⊢ᶜ$ (if b then c$₂$ else c$₃$) $⇓$ v $⊣$ (s$″$,$ρ″$))
   $→$ explicate-pred e$₁$ c$₂$ c$₃$ G $≡$ (c$₁$ , G$′$)
@@
 
-\noindent This statement of correctness is interesting because it
-requires the premise that either continuation \lstinline{c$₂$} or
-\lstinline{c$₃$} evaluates to \lstinline{v}, depending on whether the
-expression \lstinline{e$₁$} evaluates to \lstinline{true} or
-\lstinline{false}. This explains the use of Agda's
-\lstinline{if-then-else} in the second premise of
-\lstinline{explicate-pred-correct}.\footnote{The formulation of this
-lemma follows that of CompCert's \lstinline{transl\_condexpr\_prop}.}
+\noindent This statement is interesting because it requires the
+premise that either continuation \lstinline{c$₂$} or \lstinline{c$₃$}
+evaluates to \lstinline{v}, depending on whether the expression
+\lstinline{e$₁$} evaluates to \lstinline{true} or
+\lstinline{false}. This explains the 
+\lstinline{(if b then c$₂$ else c$₃$)} in the second premise.\footnote{The
+formulation of this lemma follows that of CompCert's
+\lstinline{transl\_condexpr\_prop}.}
 
 The proof of \lstinline{explicate-pred-correct} uses many of the
-monotonicity lemmas that we have already proved, plus one more.
+monotonicity lemmas that we have already proved, plus the following
+one.
 
@@
 into x86 instructions. The minor challenge is that many x86
-instructions take just two arguments and have an update-in-place
+instructions take only two arguments and have an update-in-place
 semantics. The output of Select Instructions on the running example is
@@
   \hrule
-\begin{lstlisting}[mathescape=false]
+\begin{lstlisting}[mathescape=false,basicstyle=\ttfamily\footnotesize]
     var x, y, t1, t2
@@
 
-The function \lstinline{select-inst} translation a whole program in
+The function \lstinline{select-inst} translates a whole program in
 the C-like language to x86.
@@
 \noindent The \lstinline{select-stmt} function maps each statement to
-one or more instructions. The return statement is compiled like an
-assignment to the \lstinline{rax} register. (The calling convention
-for x86 says that the return value of a function is passed in
-\lstinline{rax}.)
+one or more instructions. The \lstinline{Return} statement is compiled
+like an assignment to the \lstinline{rax} register. (The calling
+convention for x86 says that the return value of a function is passed
+in \lstinline{rax}.)
 
@@
 We compile the assignment of an equality comparison into a
-\lstinline{CmpQ} and a \lstinline{MovQ}. The \lstinline{CmpQ} always
-puts its result into \lstinline{rax}, so the \lstinline{MovQ} is
-needed to place the result into the destination.\footnote{The real
-\lstinline{cmpq} instruction is slightly stranger, involving
-the \lstinline{EFLAGS} register.}
+\lstinline{CmpQ} and a \lstinline{MovQ}. The \lstinline{CmpQ} puts its
+result into \lstinline{rax}, so the \lstinline{MovQ} is needed to
+place the result into the destination.\footnote{The real
+\lstinline{cmpq} instruction is stranger still, involving the
+\lstinline{EFLAGS} register.}
 
@@
 When the statement is an assignment, the proof uses the following
-lemma. It states that if expression \lstinline{e} evaluates to
-\lstinline{v}, then the instructions generated for \lstinline{e} put
+lemma. It states that when expression \lstinline{e} evaluates to
+\lstinline{v}, the instructions generated for \lstinline{e} put
 \lstinline{v} into the specified destination. The predicate
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[basicstyle=\ttfamily\footnotesize]
 compile-correct : (e : Exp) (s : Inputs) (v : Value)
@@
 to \citet{Aho:1986qf}, \citet{Dybvig:1990aa}, and Olivier Danvy's
-one-pass transformation into monadic normal form~\citep{Danvy:2003fk}.
+transformation into monadic normal form~\citep{Danvy:2003fk}.
 

### 2025-08-12 paper.tex
spell check

@@
 %
-There are many other pedogogical treatments of
+There are many other pedagogical treatments of
 compilation~\citep{Appel:2003fk,Sumii:2005aa,Aho:2006wb,Cooper:2011aa,Grune:2012aa}
@@
 % LocalWords:  ReadInt rax StateX EFLAGS var inst rco cmplx suc IL ie
-% LocalWords:  retB thenB pred GHC thn eval Agda's transl condexpr
-% LocalWords:  stmt arg dest regs DestOK trans
+% LocalWords:  retB thenB pred GHC thn eval Agda's transl condexpr wb
+% LocalWords:  stmt arg dest regs DestOK trans Lawall Appel Sumii cg
+% LocalWords:  Grune Isabelle Hatcliff vn Kelsey uq Birkedal movq int
+% LocalWords:  basicstyle mathescape callq subq

### 2025-08-12 paper.tex
fixes from Tianyu

@@
 where the expression $b$ is compiled using a special function
-$\mathcal{B}$ that takes two more arguments, the generated code for
-the two branches, that is, for the two continuations of $b$. Thus,
-$\mathcal{B}$ can analyze the expression $b$ to decide how to generate
-code for the $\mathsf{if}$, or whether an $\mathsf{if}$ is even
-necessary.
+$\mathcal{B}$ that takes two more arguments, the generated code $c_1$
+and $c_2$ for the two branches, that is, for the two continuations of
+$b$.  Thus, $\mathcal{B}$ can analyze the expression $b$ to decide how
+to generate code for the $\mathsf{if}$, or whether an $\mathsf{if}$ is
+even necessary.
 \[
@@
 
-The third language replaces the \lstinline{Let} bindings with
-assignment expressions but retains the structured control-flow and
+The third language replaces \lstinline{Let} bindings with assignment
+expressions but retains the structured control-flow and
 expression-oriented nature of the language. This language brings us
 closer to x86 where variables are replaced by registers and stack
-locations, all of which are mutable. The abstract syntax for
-assignment is \lstinline{(Assign $x$ $e_1$ $e_2$)} where $e_1$ is the
-right-hand side and $e_2$ is an expression to be executed after the
-assignment (reminiscent of the body of a \lstinline{Let}).  The
-\lstinline{Program} AST node consists of the number of variables and
-an expression.
+locations, which are mutable. The abstract syntax for assignment is
+\lstinline{(Assign $x$ $e_1$ $e_2$)} where $e_1$ is the right-hand
+side and $e_2$ is an expression to be executed after the assignment
+(like the body of a \lstinline{Let}).  The \lstinline{Program} AST
+node consists of the number of variables and an expression.
 
@@
 big-step semantics.
-
 \begin{lstlisting}
@@
 \end{lstlisting}
-
 \noindent Here is the rule for assignment:
@@
 \]
-
-The semantics of whole programs is given by the following relation.
-
+The semantics of programs is given by the following.
 \begin{lstlisting}
@@
 return the continuation for the \lstinline{thn} or \lstinline{els}
-branch, respectively. (This may case \lstinline{thn} or
+branch, respectively. (This may cause \lstinline{thn} or
 \lstinline{els} to become dead code, which we discuss later.)

### 2025-08-12 paper.tex
tweak

@@
 The third language replaces \lstinline{Let} bindings with assignment
-expressions but retains the structured control-flow and
-expression-oriented nature of the language. This language brings us
-closer to x86 where variables are replaced by registers and stack
-locations, which are mutable. The abstract syntax for assignment is
+expressions but retains the expression-oriented and structured
+control-flow nature of the language. This language brings us closer to
+x86 where variables are replaced by registers and stack locations,
+which are mutable. The abstract syntax for assignment is
 \lstinline{(Assign $x$ $e_1$ $e_2$)} where $e_1$ is the right-hand

### 2025-08-21 paper/paper.tex
fixes

--- 2025-08-12 paper.tex
@@
 %%
-\documentclass[sigplan,review,dvipsnames,screen,10pt]{acmart}
-%\documentclass[sigplan,screen]{acmart}
+%\documentclass[sigplan,screen,10pt]{acmart}
+\documentclass[sigplan,screen]{acmart}
 %%
@@
 %% complete the rights form.
-\setcopyright{acmlicensed}
-\copyrightyear{2018}
-\acmYear{2018}
-\acmDOI{XXXXXXX.XXXXXXX}
-%% These commands are for a PROCEEDINGS abstract or paper.
-\acmConference[OlivierFest '25]{Workshop dedicated to Olivier Danvy's 65th birthday}{October,
-  2025}{Singapore}
-%%
-%%  Uncomment \acmBooktitle if the title of the proceedings is different
-%%  from ``Proceedings of ...''!
-%%
-%%\acmBooktitle{Woodstock '18: ACM Symposium on Neural Gaze Detection,
-%%  June 03--05, 2018, Woodstock, NY}
-\acmISBN{978-1-4503-XXXX-X/2018/06}
+\setcopyright{cc}
+\setcctype{by}
+\acmDOI{10.1145/3759427.3760372}
+\acmYear{2025}
+\copyrightyear{2025}
+\acmISBN{979-8-4007-2150-2/25/10}
+\acmConference[OLIVIERFEST '25]{Proceedings of the Workshop Dedicated to Olivier Danvy on the Occasion of His 64th Birthday}{October 12--18, 2025}{Singapore, Singapore}
+\acmBooktitle{Proceedings of the Workshop Dedicated to Olivier Danvy on the Occasion of His 64th Birthday (OLIVIERFEST '25), October 12--18, 2025, Singapore, Singapore}
+\received{2025-06-18}
+\received[accepted]{2025-07-31}
 
@@
 \title{Verified Nanopasses for Compiling Conditionals}
+\titlenote{This work was partially funded by NSF award 1814460.}
 
@@
 %%
-%% \begin{CCSXML}
-%% <ccs2012>
-%%  <concept>
-%%   <concept_id>00000000.0000000.0000000</concept_id>
-%%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-%%   <concept_significance>500</concept_significance>
-%%  </concept>
-%%  <concept>
-%%   <concept_id>00000000.00000000.00000000</concept_id>
-%%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-%%   <concept_significance>300</concept_significance>
-%%  </concept>
-%%  <concept>
-%%   <concept_id>00000000.00000000.00000000</concept_id>
-%%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-%%   <concept_significance>100</concept_significance>
-%%  </concept>
-%%  <concept>
-%%   <concept_id>00000000.00000000.00000000</concept_id>
-%%   <concept_desc>Do Not Use This Code, Generate the Correct Terms for Your Paper</concept_desc>
-%%   <concept_significance>100</concept_significance>
-%%  </concept>
-%% </ccs2012>
-%% \end{CCSXML}
-
-%% \ccsdesc[500]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-%% \ccsdesc[300]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-%% \ccsdesc{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-%% \ccsdesc[100]{Do Not Use This Code~Generate the Correct Terms for Your Paper}
-
-%%
+\begin{CCSXML}
+<ccs2012>
+   <concept>
+       <concept_id>10011007.10011006.10011041</concept_id>
+       <concept_desc>Software and its engineering~Compilers</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10011007.10011074.10011099.10011692</concept_id>
+       <concept_desc>Software and its engineering~Formal software verification</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10003752.10010124.10010131.10010134</concept_id>
+       <concept_desc>Theory of computation~Operational semantics</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+   <concept>
+       <concept_id>10003752.10010124.10010138.10010142</concept_id>
+       <concept_desc>Theory of computation~Program verification</concept_desc>
+       <concept_significance>500</concept_significance>
+       </concept>
+ </ccs2012>
+\end{CCSXML}
+
+\ccsdesc[500]{Software and its engineering~Compilers}
+\ccsdesc[500]{Software and its engineering~Formal software verification}
+\ccsdesc[500]{Theory of computation~Operational semantics}
+\ccsdesc[500]{Theory of computation~Program verification}
+
 %% Keywords. The author(s) should pick words that accurately describe
@@
 case-of-case transformation is employed to optimize dispatching on
-algebraic datatypes~\citep{PEYTONJONES19983}.
+datatypes~\citep{PEYTONJONES19983}.
 
@@
 \emph{Essentials of Compilation}
-textbook~\citep{Siek:2023tr,Siek:2023ue}. While it does not include
-proofs of correctness for its compiler passes, the textbook is
-organized in a way that might facilitate the proofs of correctness.
+textbook~\citep{Siek:2023tr,Siek:2023ue}. This textbook does not
+include proofs of correctness for its compiler passes, though it is
+organized in a way that might facilitate such proofs.
 %
@@
 conditional expressions and Booleans into an x86-flavored register
-transfer language. The Agda code is available in the supplementary
-material and at the following location.
+transfer language. The Agda code is available at the following location.
 \begin{center}
-\url{https://github.com/jsiek/correct_compilers}
+\url{https://doi.org/10.5281/zenodo.16890248}
 \end{center}
@@
 
-\subsection{The C-like language}
+\subsection{The C-like Language}
 \label{sec:c-lang}
@@
 
-\subsection{x86 with variables}
+\subsection{X86 with Variables}
 
@@
 \begin{figure}[tp]
-  \hrule
+  %\hrule
 \begin{lstlisting}
@@
 \end{lstlisting}
-  \hrule
+  %\hrule
 \caption{An example source program.}
@@
 \end{lstlisting}
-  \hrule
+  %\hrule
 \caption{\small Result of Remove Complex Operands on the example.}
@@
 
-\subsubsection{Remove Complex Operands is Correct}
+\paragraph{Remove Complex Operands Is Correct}
 
@@
 \end{lstlisting}
-  \hrule
+  %\hrule
   \caption{Result of Lift Locals on the example.}
@@
 
-\subsubsection{Lift Locals is Correct}
+\paragraph{Lift Locals Is Correct}
 
@@
 \end{lstlisting}
-  \hrule
+  %\hrule
   \caption{Example after Explicate Control, where control flow
@@
 
-For the finale of the Explicate Control pass, we discuss the
-translation of the conditional \lstinline{(If e$₁$ e$₂$ e$₃$)}
-in \lstinline{explicate-pred}.  The two
-branches \lstinline{e$₂$} and \lstinline{e$₃$} are in predicate
-position, so we compile them with recursive calls to
-\lstinline{explicate-pred} to obtain \lstinline{c$₂$} and
-\lstinline{c$₃$}. However, they share the same two continuations,
-\lstinline{thn} and \lstinline{els}, so we add both of them to the CFG
-to avoid code duplication. We compile \lstinline{e$₁$} with
-\lstinline{explicate-pred}, passing \lstinline{c$₂$} and
-\lstinline{c$₃$} as the continuations.
-
-\begin{lstlisting}
+For the finale of the Explicate Control pass, we discuss the case in
+\lstinline{explicate-pred} for the conditional \lstinline{(If e$₁$ e$₂$ e$₃$)}.
+The two branches \lstinline{e$₂$} and
+\lstinline{e$₃$} are in predicate position, so we compile them with
+recursive calls to \lstinline{explicate-pred} to obtain
+\lstinline{c$₂$} and \lstinline{c$₃$}. However, they share the same
+two continuations, \lstinline{thn} and \lstinline{els}, so we add both
+of them to the CFG to avoid code duplication. We compile
+\lstinline{e$₁$} with \lstinline{explicate-pred}, passing
+\lstinline{c$₂$} and \lstinline{c$₃$} as the continuations.
+
+\begin{lstlisting}[basicstyle=\ttfamily\footnotesize]
 explicate-pred (If e$₁$ e$₂$ e$₃$) thn els =
@@
 
-\begin{lstlisting}
+\begin{lstlisting}[basicstyle=\ttfamily\footnotesize]
 explicate-pred-correct :
@@
 \begin{figure}[tp]
-  \hrule
+  %\hrule
 \begin{lstlisting}[mathescape=false,basicstyle=\ttfamily\footnotesize]
@@
 \end{lstlisting}
-  \hrule
+  %\hrule
   \caption{Example after Select Instructions, in the x86 assembly language
@@
 
-\noindent For the assignment statement we delegate to the
-\lstinline{select-assign} function, which analyzes the expression
-\lstinline{e} to determine which instruction to generate.
-We describe \lstinline{select-assign} shortly.
+\noindent For the assignment statement we delegate to
+\lstinline{select-assign}, which analyzes the expression \lstinline{e}
+to determine which instruction to generate.  We describe
+\lstinline{select-assign} shortly.
 
@@
 
-Next we look at some representative cases of
-\lstinline{select-assign}. The first parameter is the right-hand side
-of the assignment and the second parameter is the left-hand side, that
-is, the destination. For subtraction, we use the \lstinline{rax}
-register as scratch space. We first move argument \lstinline{a$₁$}
-into \lstinline{rax}, then subtract \lstinline{a$₂$} from
-\lstinline{rax}, and finally move the result into \lstinline{dest}.
+Next we look at representative cases of \lstinline{select-assign}. The
+first parameter is the right-hand side of the assignment and the
+second parameter is the left-hand side, that is, the destination. For
+subtraction, we use the \lstinline{rax} register as scratch space. We
+first move argument \lstinline{a$₁$} into \lstinline{rax}, then
+subtract \lstinline{a$₂$} from \lstinline{rax}, and finally move the
+result into \lstinline{dest}.
 
@@
 
-\subsubsection{Select Instructions is Correct}
+\paragraph{Select Instructions Is Correct}
 
@@
 
-The correctness theorem delegates to the correctness lemma for
-\lstinline{select-stmt}, which says that if statement \lstinline{c}
-evaluates to \lstinline{v}, then the instructions generated for
-\lstinline{c} put the same value \lstinline{v} into \lstinline{rax}.
-The list of x86 instruction blocks is obtained by mapping
-\lstinline{select-stmt} over the control-flow graph \lstinline{G}.
+The correctness theorem delegates to a lemma about the
+\lstinline{select-stmt} function, which says that if statement
+\lstinline{c} evaluates to \lstinline{v}, then the instructions
+generated for \lstinline{c} put the same value \lstinline{v} into
+\lstinline{rax}.  The list of x86 instruction blocks is obtained by
+mapping \lstinline{select-stmt} over the control-flow graph
+\lstinline{G}.
 
@@
 
-\subsection{The Compiler is Correct}
+\subsection{The Compiler Is Correct}
 
@@
 transformation into monadic normal form~\citep{Danvy:2003fk}.
+
+\section{Data-Availability Statement}
+
+The Agda files the define the compiler and proof of correctness are
+available from \citet{Siek:2025aa}.
+
+\balance
 
